---
title: "SNA lab 1"
author: "Alexandra Rottenkolber, based on a lab written by José Luis Estévez Navarro"
date: "2026-01-23"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-expand: 1
    toc-title: Contents
    number-sections: true
    number-offset: -1
    css: style.css
    embed-resources: true
    theme: cosmo
editor_options: 
  chunk_output_type: console
---

# Content of this lab: Introduction

1. Data formats and packages for social network analysis in R
2. Reading and exploring network data
3. Network visualizations
4. Elementary transformation: dichotomization, transposition, and symmetrization


```{r setup 1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Clean the environment
rm(list=ls())

library(sna) # networks package
library(igraph) # networks package 
library(tidygraph) # analysis of network data, syntax similiar to tidyverse (more here https://schochastics.github.io/tidynetAnaR/)
# beware of masked functions
library(gtools) 
library(ggplot2) # nice plots
library(ggraph) # plotting of network data, syntax similiar to ggplot (more here https://schochastics.github.io/tidynetAnaR/)
library(dplyr) # data manipulation
library(data.table) # data handling
library(gridExtra)

# Sometimes convenient to detach packages
# detach(package:sna,unload = TRUE)

# Or call functions by specifying the package they are from
# sna::betweenness()
# igraph::betweenness()

options(scipen=999) # set notation option

#knitr::opts_knit$set(root.dir = "YOUR PATH")
knitr::opts_knit$set(root.dir = "/Users/alexandrarottenkolber/Documents/03_PhD_at_LiU/08_Sessions_by_me/SNA/2026_Lab_sessions_1_4/session_1")

```

# Data formats and packages for social network analysis in R

## Data formats

```{r data formats 1}
# Vectors, matrices, and data frames
# A VECTOR is collection of elements (often of the same type)
v <- c(0,1,2,3,4,5)
v
```

```{r data formats 2}
# If elements of different class (numeric, logical, characters), R uses the less exigent class for the vector
v2 <- c(1,2,3,'a','b',TRUE,FALSE,NaN)
v2 
```

```{r data formats 3}
# Notice that numbers (ar logical operators) were turned into "character" 
class(v)
class(v2)
```

```{r operations}
v+1 # scalars are applied to all elements of a vector
v*rev(v) # in vectors of the same length, operations are performed element to element 
v*c(1,2) # otherwise, R repeats the shorter vector
```

```{r dataframes}
# We are accustomed to see vectors as columns in a DATA FRAME
data <- mtcars
class(data)
#View(data)
head(data,20) # show the first 20 rows of the data frame

data$mpg # The first column is a vector
data$cyl # The second column is another vector
# In short, a data frame is simply a collection of vectors. 
```

```{r matrices 1}
# In SNA, the basic element is the MATRIX
# Whereas a vector is uni-dimensional, a matrix is two-dimensional
mtx <- matrix(c(0,1,1,
                1,0,0,
                2,1,0),
              nrow=3,ncol=3,byrow=TRUE)
mtx
```

```{r matrices 2}
class(mtx) # An array is a n-dimensional object (you can create objects of 3, 4,... dimensions)
dim(mtx) # Shows the number of elements in each dimension (a 3x3 matrix in this case)

# Notice you would normally name the dimensions 
dimnames(mtx) <- list(c('Amy','Bjorn','Carl'),c('Amy','Bjorn','Carl'))
mtx
```

```{r matrices 3}
# Each element of the matrix can be accessed using the position (or the names if available)
mtx[1,2] # first row, second column
mtx['Carl','Amy'] # third row, first column
```

```{r matrices 4}
# You can replace elements too
mtx[3,1] <- 0 # This overwrites the element corresponding to 3rd row, 1st column
mtx
```

```{r matrices 5}
# In SNA, one important function for matrices is the diagonal
diag(mtx) # The elements in positions (1,1), (2,2), (3,3),... (n,n)
diag(mtx) <- NA # Often sent to missing data. Done because having it set to zero would mean that there are no self-loops, 
# while NA would mean that the ties itself cannot exist per definition, and are just excluded
# Important to note that missing is different from the zero!
mtx
```

```{r matrices 6}
# Visualize the adjacency matrix
mtx |> 
  reshape2::melt() |> 
ggplot2::ggplot(ggplot2::aes(x=Var1,y=Var2,fill=value |> as.factor()))+
  ggplot2::geom_tile()+
  ggplot2::labs(title='Adjacency matrix', x=NULL, y=NULL)+
  ggplot2::scale_fill_manual(values=c('white','steelblue', "lightgray"), name='Contact', labels=c('No','Yes', NA))+
  ggplot2::theme_minimal()
```



```{r edgelist}
# What we have now could be an actual network
# Say the values indicate whether person i (in the rows) respects person j (in the columns)
# How would you interpret this network?

# Matrices are heavy objects. Often network data is stored as edge lists
edgelist <- data.frame(person1=c('Amy','Amy','Bjorn','Carl'),
                       person2=c('Bjorn','Carl','Amy','Bjorn'))
edgelist
# No worry about this, we will use some existing packages to perform the transformation
```

## R packages for social network analysis

There are plenty of packages tailored to perform different types of SNA in R. Among these,

  + "sna" 
  + "igraph"
  + "network"
  + "tidygraph" https://tidygraph.data-imaginist.com/ 
  + "ggraph"  for visualisations (uses ggplot2)
  + "statnet"  meta-package; herein you have "ergm", "tergm", etc.
  + "RSiena"  for Stochastic Actor-Oriented Models (SAOMs)
  + "intergraph"  for conversion between network data objects (igraph, network, matrix, edge list, etc.)
  + "influenceR"  for measures of structural importance (effective network size, constraint, bridging)
  + "egor"  for ego-nets (uses igraph objects)
  + "netseg"  for assortativity patterns, homophily, segregation, within-group mixing, etc.
  + "isnar"
  + "netropy" 
  + "networkdata" a large sample of networks in igraph format (to install networkdata, see: https://github.com/schochastics/networkdata)
  + etc.

To install packages, type the name of the package within (don't forget quotation marks): install.packages(""),
install.packages(c('sna','igraph')),
install.packages("isnar")

To use them, type the name of the package within: library()

library(sna)

detach("package:sna") --> detach is useful when functions from different packages overlap, so the first function is masked

The most widely used R packages for (simple) network analysis are called `sna` and `igraph`. 

## Data handling 

```{r data handling 1}

# reading in and exploring network data, an example 

# Information on the data we will use
#browseURL("https://www.stats.ox.ac.uk/~snijders/siena/s50_data.htm")

edges_t1 <- read.csv("./s50_data/s50-network1.dat", header = FALSE, sep = " ") # adjacency matrix of the student network
edges_t1$V1 <- NULL # remove index

node_sport <- read.csv("./s50_data/s50-sport.dat", header = FALSE, sep = " ") # How much sport people do 
node_sport$V1 <- NULL # remove index
names(node_sport) <- c("t1", "t2", "t3") # set column names
node_sport["node_id"] <- colnames(edges_t1)

node_alc <- read.csv("./s50_data/s50-alcohol.dat", header = FALSE, sep = " ") # How much alcohol people drink
node_alc$V1 <- NULL # remove index
names(node_alc) <- c("t1", "t2", "t3") # set column names
node_alc["node_id"] <- colnames(edges_t1)
```

```{r data handling 2}

# Data exploration
class(node_sport); class(node_alc); class(edges_t1)
dim(node_sport); dim(node_alc); dim(edges_t1)

head(node_sport) # this is a standard data set, each row represents a person (N = 50)
# For your own analysis it it always good to check for duplicates
# length(unique(nodes$ID)) == nrow(nodes))
colnames(node_sport)

head(edges_t1) # This is an adjacency matrix (student -> reported friend)
colnames(edges_t1)
# If we had an edge list, it makes sense check whether there are any duplicated ties (any(duplicated(edge_list))

```

## Graph objects

To create a graph object, SNA works with matrices as inputs, while this is not the case for IGRAPH. IGRAPH requires igraph objects as inputs. You can find information on the basic IGRAPH datatypes [here](https://igraph.org/c/doc/igraph-Basic.html).

```{r data handling 3}

# Conversion to igraph graph
mtx <- as.matrix(edges_t1)
grph <- graph_from_adjacency_matrix(mtx) # igraph functions take matrices as inputs
length(as_edgelist(grph)) # 226 ties
length(isolates(mtx)) # number of isolates
```

# Visualisation

## Simple visualisation
```{r visualisation 1}
# visualise graph
sna::gplot(mtx,displaylabels = TRUE)


# Using the igraph library

#plot.igraph(mtx) # this will throw an error 
grph |> 
  igraph::plot.igraph(
            vertex.label='', vertex.size=4,
            edge.color='darkgrey', edge.arrow.size=.2,
            layout=layout_with_kk(grph), # other options: layout_nicely, layout_with_fr (<-- this one is furchterman reingold)
            main='Friendship network')


# Using the ggraph library
grph |> 
  ggraph::ggraph(layout = "kk") +
  geom_edge_link() +
  geom_node_label(aes(label=name)) +
  theme_void() +
  ggtitle("Network visualisation with ggraph")
```

## Visualisation of node attributes

```{r visualisation 2}
# Addition of nodes' attributes (amount of sport and alcohol consumption)
# Be careful to correctly match the igraph object with attribute data
V(grph)$name[1:10] # This is the order in which nodes appear in the igraph object
node_sport$node_id[1:10] # This - the order in the other element
# Use the match function
node_sport[match(V(grph)$name,node_sport$node_id),]$t1 # usually nodes and edges are not aligned in the same order => make sure you assign the right attributes to the right node 
node_alc[match(V(grph)$name,node_alc$node_id),]$t1 

V(grph)$sport <- node_sport[match(V(grph)$name,node_sport$node_id),]$t1
V(grph)$alc_con <- node_alc[match(V(grph)$name,node_alc$node_id),]$t1

# Final inspection of the igraph object
grph
```

```{r visualisation 3}

# Remember nodes attributes can be visualized
# -- Sport
plot.igraph(grph,
            vertex.label='',vertex.size=4,vertex.color=ifelse(V(grph)$sport == 2,'royalblue','orange'), # 2 means regular sport
            edge.color='darkgrey',edge.arrow.size=.2,
            layout=layout_with_kk(grph),
            main='Friendship network with level of sport \nblue: regular, orange: not regular')


legend(
  x = 1.2, 
  y = 0.2,
  xpd = TRUE,
  legend = c("Regular", "Not regular"),
  title = "Sport",
  pch = 21,
  pt.bg = c("royalblue", "orange"),
  pt.cex = 1.5,
  bty = "n",
  inset = 0.02
)
  


```

```{r visualisation 4}

# -- Alcohol consumption
# Create a palette
unique(V(grph)$alc_con)
pal <- c("royalblue", "lightblue", "lightgreen", "orange", "red") # Alcohol: 1 (non), 2 (once or twice a year), 3 (once a month), 4 (once a week) and 5 (more than once a week).

# Get each node's consumption behaviour
nodeCon <- V(grph)$alc_con[!duplicated(V(grph)$name)]

# Assign to the vertices as a color attribute
V(grph)$alc_color <- pal[nodeCon]

plot.igraph(grph,
            vertex.label='', 
            vertex.size=4,
            vertex.color=V(grph)$alc_color, # 2 means regular sport
            edge.color='darkgrey',
            edge.arrow.size=.2,
            layout=layout_with_kk(grph),
            main='Friendship network with alcohol consumption \nblue (non), lightblue (once or twice a year), green (once a month),\norange (once a week) and red (more than once a week)')

legend(
  x = 1.2, 
  y = 0.5,
  xpd = TRUE,
  legend = c(
    "non",
    "once or twice\na year",
    "once a month",
    "once a week",
    "more than once\na week"
  ),
  title = "Alcohol consumption",
  pch = 21,
  pt.bg = pal,
  pt.cex = 1.5,
  bty = "n",
  inset = 0.02
)
```

# Transformations

## Dichotomization (recoding)

```{r transformations 1}

# Useful when valued or weighted matrices, but we aim for a simple analysis (logistic regression)
# Image, respondents were asked to assess their relationship with others on a scale 1 to 5

w.mtx <- matrix(c(0,5,2,1,2,
                  4,0,0,3,2,
                  2,2,0,4,3,
                  1,2,5,0,3,
                  1,1,5,4,0),
                nrow=5,ncol=5,byrow=TRUE,
                dimnames = list(c('A','B','C','D','E'),c('A','B','C','D','E')))
w.mtx

# igraph object
w.grph <- graph_from_adjacency_matrix(w.mtx,weighted = TRUE)
w.grph
plot.igraph(w.grph) # a fully connected network
plot(w.grph,
     edge.width=E(w.grph)$weight, 
     main="weights visible", 
     edge.curved=c(rep(0.1, 19))
     ) # With some ties stronger than others

# You decide that at least a 4 needs to be given for a tie to exist
# Let's keep only strong ties: 4 OR 5
dich.mtx <- w.mtx
dich.mtx[dich.mtx %in% 0:3] <- 0
dich.mtx[dich.mtx %in% 4:5] <- 1
dich.mtx

dich.grph <- graph_from_adjacency_matrix(dich.mtx)
dich.grph
plot(dich.grph,
     main='dichotomized version',  
     edge.curved=c(rep(0.25, ecount(dich.grph)))
     )
```

## Transposition (shifting rows for columns)

```{r transformations 2}

# Reversing the direction of the ties
# Useful depending on the data collection procedure
(dich.mtx.t <- t(dich.mtx))
dich.grph.t <- graph_from_adjacency_matrix(dich.mtx.t)
plot(dich.grph.t,main='transposed version', 
     edge.curved=c(rep(0.25, ecount(dich.grph))))
```

## Symmetrization (from directed to undirected network)

```{r transformations 3}
(sym.mtx.weak <- symmetrize(dich.mtx,rule='weak')) # asymmetric ties kept (max-symmetrizing)
(sym.mtx.stro <- symmetrize(dich.mtx,rule='strong')) # only mutual ties are kept (min-symmetrizing)


# Visualize the adjacency matrix
plt1 <- sym.mtx.weak |> 
  reshape2::melt() |> 
ggplot2::ggplot(ggplot2::aes(x=Var1,y=Var2,fill=value |> as.factor()))+
  ggplot2::geom_tile()+
  ggplot2::labs(title='Visualise weak symmerization', x=NULL, y=NULL)+
  ggplot2::scale_fill_manual(values=c('white','steelblue', "lightgray"), name='Contact', labels=c('No','Yes', NA))+
  ggplot2::theme_minimal()

# Visualize the adjacency matrix
plt2 <- sym.mtx.stro |> 
  reshape2::melt() |> 
ggplot2::ggplot(ggplot2::aes(x=Var1,y=Var2,fill=value |> as.factor()))+
  ggplot2::geom_tile()+
  ggplot2::labs(title='Visualise strong symmetrization', x=NULL, y=NULL)+
  ggplot2::scale_fill_manual(values=c('white','steelblue', "lightgray"), name='Contact', labels=c('No','Yes', NA))+
  ggplot2::theme_minimal()

gridExtra::grid.arrange(plt1, plt2, ncol=2)

# The function removes the nodes' names
dimnames(sym.mtx.weak) <- list(c('A','B','C','D','E'),c('A','B','C','D','E'))
dimnames(sym.mtx.stro) <- list(c('A','B','C','D','E'),c('A','B','C','D','E'))

# Convert into igraph objects
sym.grph.weak <- graph_from_adjacency_matrix(sym.mtx.weak,mode='undirected') # notice the argument: mode="undirected"
sym.grph.stro <- graph_from_adjacency_matrix(sym.mtx.stro,mode='undirected') # notice the argument: mode="undirected"

par(mfrow=c(1,2)) # Two plots side to side
plot.igraph(sym.grph.weak,main='Max-symmetrized')
plot.igraph(sym.grph.stro,main='Min-symmetrized')
par(mfrow=c(1,1))
```












