---
title: "Lab 1"
author: " $ patricia  givort cruz cabral"
date: "2026-08-23"
format:
  pdf:
    number-sections: false
    fig-cap-location: top
    tbl-cap-location: top
    fig-numbering: true
    tbl-numbering: true
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
---


```{r setup 1, inclue=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(data.table)  # Importing and manipulating data
library(glmnet)      # Ridge/lasso regression
library(ggplot2)     # Plotting
library(mlbench)     # Example data used below
library(caret)       # Cross-validation and model evaluation
library(splines)     # Natural splines: ns()
library(ggeffects)   # Model-based predictions: ggpredict()
setwd('C:/Users/paty_/OneDrive/Documentos/LAB machine learning/Lab1/Lab1')

load("")



```

```{r,inclue=FALSE }
#| echo: false

# --------------------------------------------------------
# High-dimensional data example
# --------------------------------------------------------

# Objective: predict sentiment based on language used in IMDB reviews

# Import (pre-processed) IMDB data.
# Keep this R file and imdb_data.csv in the same folder.
trumpbernie <- fread(file = "trumpbernie.csv")

# Inspect
trumpbernie[1:10,1:10]
dim(trumpbernie) # High-dimensional? Yes.

# Format the outcome as a factor for classification with caret
trumpbernie[, trump_tweet := as.factor(trump_tweet)]



colnames(trumpbernie)
# Fit ridge regression
# Assess a range of lambda values through (5-fold) cross-validation
X <- trumpbernie[,-c('id','trump_tweet'),with=F] # Exclude response and id columns.
X <- as.matrix(X) # Make input data into a matrix

# Important: standardize=TRUE makes glmnet standardize X internally when
# fitting/penalizing the model. However, coef() later returns coefficients
# transformed back to the original X scale.

set.seed(12345)
cvglmnet <- cv.glmnet(x = X, 
                      y = imdb$sentiment,
                      nfolds = 5,              # Number CV-folds
                      standardize = TRUE,      # Standardize X internally when fitting
                      family='binomial',       # Outcome binary --> logit/binomial
                      alpha=0,                 # alpha=0 --> ridge
                      type.measure = 'class')  # CV criterion: misclassification error

# Which lambda gave the lowest cross-validated error?
cvglmnet
cvglmnet$lambda.min

# Convert the lowest misclassification error to accuracy
1 - min(cvglmnet$cvm)

plot(cvglmnet, sign.lambda = 1) 

# From plot:
# i) Large penalties can worsen performance by increasing bias.
# ii) Note that lambda is shown on a log scale.



```



# Introduction

> 




## 1. Explain your expectation!

> 






## 2. Friendship  expectation


>


## 3. Control mechnisms

> 





## 4. Hierarchy index


> 






## 5. Descriptive comparison

```{r, inclue=FALSE}
calc_net_stats <- function(adj, gender_vec){

  adj <- as.matrix(adj)
  diag(adj) <- 0
  adj[is.na(adj)] <- 0

  g <- igraph::graph_from_adjacency_matrix(adj, mode = "directed", diag = FALSE)

  dens  <- igraph::edge_density(g, loops = FALSE)
  recip <- igraph::reciprocity(g)

  trans <- igraph::transitivity(g, type = "global")
  if (is.nan(trans)) trans <- NA_real_

  gender_f <- as.factor(gender_vec)
  igraph::V(g)$gender <- gender_f
  gender_codes <- as.integer(igraph::V(g)$gender)
  homoph <- igraph::assortativity_nominal(g, types = gender_codes, directed = TRUE)

  hier <- igraph::centr_degree(g, mode = "in", normalized = TRUE)$centralization

  c(
    density = dens,
    reciprocity = recip,
    transitivity = trans,
    gender_homophily = homoph,
    indegree_centralization = hier
  )
}

nets <- list(
  friendship  = friends,
  liking      = liking,
  popularity  = popular
)
results <- lapply(nets, calc_net_stats, gender_vec = gender) |>
  dplyr::bind_rows(.id = "network")

results

```

```{r}
#| label: tbl-network-comparison
#| tbl-cap: "Comparison of network properties across the three directed nomination networks."
#| echo: false

results_fmt <- results |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))

knitr::kable(results_fmt, booktabs = TRUE,
             col.names = c("Network", "Density", "Reciprocity", "Transitivity",
                           "Gender homophily", "Indegree centralization"))

```



> In @tbl-network-comparison, I compare the three directed nomination networks, friendship, liking, and popularity, in terms of density, reciprocity, transitivity, gender homophily, and my hierarchy index from Exercise 4, indegree centralization. Density indicates how common nominations are in each network. The liking network is the most dense, with a value of 0.330, followed by friendship with 0.195, while popularity is much sparser at 0.041. This suggests that liking nominations are relatively frequent, friendship nominations are more selective, and popularity nominations are the most restrictive form of recognition in this group. Reciprocity is relatively high in friendship, 0.443, and liking, 0.373, indicating a substantial tendency for mutual nominations in these networks. In contrast, reciprocity in the popularity network is 0.000. This is consistent with the interpretation of popularity as a hierarchical status construct rather than a mutual relationship. Being recognized as popular does not require reciprocal acknowledgment.

> Transitivity is highest in the liking network, 0.636, and also substantial in friendship, 0.497, while much lower in popularity, 0.219. This means that liking and friendship networks show stronger tendencies toward triadic closure, where friends or liked peers of friends are also connected. Popularity appears less structured by cohesive group processes and more by individual visibility.

> Gender homophily is strongest in the friendship network, 0.504, moderate in popularity, 0.241, and lowest in liking, 0.126. This suggests that friendships are more segregated by gender, while liking nominations are more likely to cross gender boundaries. Popularity again lies in between, possibly reflecting both social grouping and broader recognition processes.

> Finally, using indegree centralization as my hierarchy index from Exercise 4, I find relatively similar values across networks: 0.127 for friendship, 0.134 for liking, and 0.138 for popularity. Since indegree centralization captures how unequal the distribution of nominations is, these results indicate that all three networks show some degree of differentiation around highly nominated individuals. Popularity is slightly more centralized, which aligns with the idea that status hierarchies are organized around a small elite. However, the differences are modest, suggesting that hierarchy is present in all three networks rather than being exclusive to popularity.










## 6. fit an exponential random graph




```{r q6-try-mcmle, include=FALSE}
rhs <- ~ edges + mutual + nodematch("gender") +
  gwesp(0.7, fixed = TRUE) + twopath + gwidegree(0.7, fixed = TRUE)

ctrl_sa <- control.ergm(
  main.method = "Stochastic-Approximation",
  SA.initial_gain = 0.02,
  SA.burnin = 2000,
  seed = 123
)

m_friend <- ergm(update(friendnet ~ 1, rhs), control = ctrl_sa)
m_like   <- ergm(update(likenet   ~ 1, rhs), control = ctrl_sa)
m_pop    <- ergm(update(popnet    ~ 1, rhs), control = ctrl_sa)

```





```{r, include=FALSE}

# 1) network objects
friendnet <- network(friends,  directed = TRUE, matrix.type = "adjacency")
likenet   <- network(liking,   directed = TRUE, matrix.type = "adjacency")
popnet    <- network(popular,  directed = TRUE, matrix.type = "adjacency")

# 2) gender attribute
gender_clean <- as.character(gender_lab)
friendnet %v% "gender" <- gender_clean
likenet   %v% "gender" <- gender_clean
popnet    %v% "gender" <- gender_clean

# 3) model RHS (required specification)
rhs <- ~ edges + mutual + nodematch("gender") +
  gwesp(0.7, fixed = TRUE) + twopath + gwidegree(0.7, fixed = TRUE)

# 4) control (use Stochastic Approximation)
ctrl <- control.ergm(
  main.method = "Stochastic-Approximation",
  SA.initial_gain = 0.05,
  SA.burnin = 2000,
  seed = 123
)

# 5) fit models
m_friend <- ergm(update(friendnet ~ 1, rhs), control = ctrl)
m_like   <- ergm(update(likenet   ~ 1, rhs), control = ctrl)
m_pop    <- ergm(update(popnet    ~ 1, rhs), control = ctrl)

#  smaller decay for gwidegree 
rhs_like03 <- ~ edges + mutual + nodematch("gender") +
  gwesp(0.7, fixed = TRUE) + twopath + gwidegree(0.3, fixed = TRUE)

m_like03 <- try(ergm(update(likenet ~ 1, rhs_like03), control = ctrl), silent = TRUE)
if (!inherits(m_like03, "try-error")) {
  m_like <- m_like03
}

# 6) extract coefs safely, and fix mutual in popularity  -Inf
extract_cells <- function(m, model_name){
  co <- as.matrix(summary(m)$coefficients)

  est <- co[, "Estimate"]
  se  <- co[, "Std. Error"]

  cell <- ifelse(
    is.infinite(est) | abs(est) > 1e6,
    "fixed at -Inf",
    sprintf("%.3f (%.3f)", est, se)
  )

  tibble(
    term  = rownames(co),
    model = model_name,
    cell  = cell
  )
}

tbl_q6 <- bind_rows(
  extract_cells(m_friend, "Friendship"),
  extract_cells(m_like,   "Liking"),
  extract_cells(m_pop,    "Popularity")
) %>%
  mutate(Term = case_when(
    term == "edges" ~ "Edges",
    term == "mutual" ~ "Mutual",
    term == "nodematch.gender" ~ "Gender homophily (nodematch)",
    grepl("^gwesp", term) ~ "Transitive closure (GWESP 0.7)",
    term == "twopath" ~ "Two-path",
    grepl("^gwideg|^gwidegree", term) ~ "Preferential attachment (GW indegree 0.7)",
    TRUE ~ term
  )) %>%
  select(Term, model, cell) %>%
  pivot_wider(names_from = model, values_from = cell) %>%
  slice(match(
    c("Edges","Mutual","Gender homophily (nodematch)","Transitive closure (GWESP 0.7)","Two-path","Preferential attachment (GW indegree 0.7)"),
    Term
  ))

```

```{r}

#| label: tbl-q6-ergm-comparison
#| tbl-cap: "ERGM results for friendship, liking, and popularity networks. Entries show estimate (SE). Models were estimated using stochastic approximation due to degeneracy issues in MPLE or MCMLE."
#| echo: false
#| message: false
#| warning: false

knitr::kable(tbl_q6, booktabs = TRUE)
```

> Across the three directed nomination networks, the ERGM results reported in Table @tbl-q6-ergm-comparison show both common structural tendencies and important differences in magnitude and stability. Because some models exhibited degeneracy and mixing problems, especially in the liking and popularity networks, we followed the professor’s instructions and estimated the models using the Stochastic Approximation algorithm with a reduced initial gain parameter to improve stability. Even so, some coefficients remain unstable or fixed, and these must be interpreted with caution. First, the edges term is negative in all three networks, indicating overall sparsity. The estimates are −3.705 (0.353) for friendship, −51674.561 (10.555) for liking, and −2.970 (0.430) for popularity. For friendship and popularity, the negative coefficients clearly reflect low baseline tie probabilities. The extremely large negative estimate in the liking network, however, indicates estimation instability rather than a substantively meaningful magnitude. Reciprocity (mutual) varies strongly across networks. In the friendship network, the estimate is positive but modest, 0.201 (0.287), suggesting limited additional reciprocation once other structural mechanisms are controlled. In the liking network, the mutual parameter is extremely large and positive, 75.529 (0.273), again indicating instability under the required specification. In the popularity network, the mutual parameter is fixed at −Inf, meaning that mutual nominations are at their minimum attainable value in the data. This implies that reciprocity is structurally absent in the popularity network. The model therefore penalizes reciprocated ties infinitely strongly. This result is substantively coherent: popularity nominations represent reputational recognition rather than mutual exchange. Gender homophily (nodematch) is positive in friendship, 0.314 (0.074), and popularity, 0.579 (0.304), indicating that same-gender nominations are more likely than cross-gender nominations in these networks. In the liking network, the estimate is −37.820 (0.089), which is again a sign of instability rather than clear evidence of reversed homophily. Overall, we find robust evidence of gender segmentation, especially in friendship and popularity, consistent with well-established adolescent homophily patterns. Transitive closure, captured by the GWESP term, is positive in all networks: 1.738 (0.131) for friendship, 2.330 (0.188) for liking, and 1.876 (0.440) for popularity. This indicates strong clustering tendencies across networks. Nominations tend to close triangles rather than remain independent, which aligns with expectations about peer group cohesion. This effect appears relatively stable across the three models. The twopath term is negative in friendship, −0.169 (0.031), and popularity, −0.304 (0.150), helping to stabilize the model by counterbalancing excessive clustering. In the liking network, the positive and large estimate, 16.974 (0.085), again signals instability. As discussed in class, the twopath term is included to reduce degeneracy-related problems, and its presence is methodologically important even when interpretation is limited.

>Preferential attachment, measured by the GW indegree term, differs substantially. In the friendship network, the estimate is positive, 3.271 (1.556), suggesting some concentration of incoming nominations on a smaller set of individuals. This provides moderate evidence of a Matthew effect in friendship. In the liking network, the parameter is fixed at −Inf, indicating that cumulative advantage cannot be stably estimated under the imposed specification. In the popularity network, the estimate is −1.102 (0.640), suggesting no clear evidence for preferential attachment once other structural mechanisms are controlled.

>Comparing only what can meaningfully be compared across models, we observe robust evidence of transitive closure and gender homophily, especially in friendship and popularity. Reciprocity is present in friendship but structurally absent in popularity. Preferential attachment appears moderate in friendship but is not clearly supported in popularity and is unstable in liking. These results partially align with our expectations: while we anticipated the strongest Matthew effect in popularity, the structural absence of reciprocity and estimation constraints suggest that popularity nominations operate under different structural rules than liking or friendship. As emphasized by the professor, some of these results may reflect unconverged or partially unstable models, and therefore substantive interpretation must remain cautious.



## 7.  interpret these results


>In light of our expectations from Exercises 1, 2, and 3, the ERGM results in Table @tbl-q6-ergm-comparison provide mixed support and also highlight important limitations due to estimation instability in parts of the liking and popularity models. Following the professor’s guidance, we relied on Stochastic Approximation controls to obtain estimates under degeneracy and mixing issues, and we therefore interpret the results cautiously, focusing mainly on patterns that are stable and substantively comparable across networks.


>Regarding our main expectation in Exercise 1, we expected preferential attachment to be stronger in popularity than in liking. The results do not clearly support this. In the popularity network, the GW indegree estimate is negative, −1.102 (0.640), which does not indicate the expected cumulative advantage pattern once other dependencies are controlled. In contrast, the liking network shows a non interpretable outcome for GW indegree, reported as fixed at −Inf, together with other extreme coefficients, indicating that preferential attachment is not reliably estimable under the required specification for this dataset. Because the liking model is unstable, we cannot make a clean popularity versus liking comparison for preferential attachment in the way we initially expected.


>For Exercise 2, we expected the Matthew effect in friendship to be intermediate between liking and popularity. What we can say with confidence is that friendship does show a positive GW indegree effect, 3.271 (1.556), suggesting some concentration of incoming nominations in friendship, consistent with a moderate cumulative advantage dynamic. However, because liking is not stably estimated and popularity shows a negative coefficient, we cannot place friendship “between” the two in a meaningful, comparable way. The most defensible interpretation is simply that friendship displays evidence of indegree concentration, while the other two networks do not provide stable, directly comparable evidence for the same mechanism under this specification.


>Exercise 3 focused on the need to control for fundamental structural mechanisms, and here the results align strongly with our expectations. Gender homophily is positive in friendship, 0.314 (0.074), and in popularity, 0.579 (0.304), consistent with the idea that same gender nominations are more likely and that failing to control for this would bias interpretation of other effects. Transitive closure is also consistently positive across networks, with GWESP estimates of 1.738 (0.131) for friendship, 2.330 (0.188) for liking, and 1.876 (0.440) for popularity, supporting our expectation that clustering and local cohesion are central features of adolescent nomination networks. Reciprocity behaves as expected in the sense that it is at least feasible in friendship, 0.201 (0.287), but it is structurally absent in popularity, where the mutual term is fixed at −Inf, indicating that reciprocated popularity nominations essentially do not occur in the observed data. This directly supports our argument that reciprocity must be modelled when present, but also that some nomination types, such as popularity, can be fundamentally asymmetric by construction.








## 8. optional




## 9. a brief assessment of goodness of fit

```{r}
#| label: fit-q9-gof
#| echo: false
#| message: false
#| warning: false

set.seed(888)

gof_friend <- gof(m_friend, control = control.gof.ergm(seed = 888, nsim = 200))
gof_like   <- gof(m_like,   control = control.gof.ergm(seed = 888, nsim = 200))
gof_pop    <- gof(m_pop,    control = control.gof.ergm(seed = 888, nsim = 200))


```

```{r}
#| label: fig-q9-gof-friend
#| fig-cap: "Goodness of fit for the friendship ERGM. The black line is the observed network, boxplots show simulated networks (central 95% intervals)."
#| echo: false
#| message: false
#| warning: false

plot(gof_friend)


```


```{r}
#| label: fig-q9-gof-like
#| fig-cap: "Goodness of fit for the liking ERGM. Interpretation should be cautious due to estimation instability noted in Exercise 6."
#| echo: false
#| message: false
#| warning: false

plot(gof_like)
```
```{r}
#| label: fig-q9-gof-pop
#| fig-cap: "Goodness of fit for the popularity ERGM. Interpretation should be cautious due to structural constraints and fixed parameters in Exercise 6."
#| echo: false
#| message: false
#| warning: false

plot(gof_pop)

```


```{r}
#| label: tbl-q9-gof-pvalues
#| echo: false
#| message: false
#| warning: false

gof_friend
gof_like
gof_pop

```


## 10.  model but only to the friendship network

```{r}

```



```{r}

```















## 11. Report converged results




















## 12.  collaborated 

























```{r}

```



```{r}

```
.
