---
title: "Final project "
author: " Themis Oluwatomole Ejegreh $ patricia  givort cruz cabral"
date: "2026-02-28"
format:
  pdf:
    number-sections: false
    fig-cap-location: top
    tbl-cap-location: top
    fig-numbering: true
    tbl-numbering: true
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
---

## Introduction

This report follows the required steps for a longitudinal social network analysis using stochastic actor-oriented models (SAOMs) in **RSiena**. We analyze directed *“best friends”* nominations collected in Dutch secondary school classrooms across multiple waves (waves **V, X, Y**).
**Data source.** The classroom network and pupil attributes come from the dataset collected by Andrea Knecht as part of her dissertation research (early adolescence, 2003/04) 

 We examine whether (i) sex homophily shapes best-friend nominations (dyadic hypothesis), (ii) triadic closure/transitivity drives clustering (triadic hypothesis), and (iii) a node-level attribute (sex-based activity) predicts how many best-friend nominations pupils send (ego/actor hypothesis). We also include standard endogenous network controls (e.g., reciprocity and degree-related effects) and evaluate model fit using goodness-of-fit (GoF) simulations.


## 1. Network Descriptives

```{r}
#| label: setup-1
#| include: false
#| echo: false
#| warning: false
#| message: false
library(haven)
library(tidyverse)
library(igraph)
library(ggplot2)
library(dplyr)
library(tibble)
library(knitr)


# Adjust the path below to where your .sav files are stored.
# Tip: if the .sav files are in the same folder as this .qmd, use data_dir <- "."
data_dir <- "."

waveV_path <- file.path(data_dir, "PupilsWaveV.sav")
waveW_path <- file.path(data_dir, "PupilsWaveW_geanonimiseerd.sav")
waveX_path <- file.path(data_dir, "PupilsWaveX.sav")
waveY_path <- file.path(data_dir, "PupilsWaveY.sav")

waveV <- read_sav(waveV_path)
waveW <- read_sav(waveW_path)
waveX <- read_sav(waveX_path)
waveY <- read_sav(waveY_path)
```

```{r}
#| label: select-class
#| echo: false
class_id <- "09f"

classV <- subset(waveV, schoolnr == class_id)
classX <- subset(waveX, schoolnr == class_id)
classY <- subset(waveY, schoolnr == class_id)
```


> We focus on classroom **`r class_id`** (variable `schoolnr`). This classroom was selected because it provides at least three observed waves and shows sufficient stability (composition and tie Jaccard) to support SAOM estimation while still exhibiting meaningful change over time.


<!--
ORIGINAL (Themis):
create_network <- function(df, friend_cols){

  ids <- df$namenr      # student IDs
  n <- length(ids)
  mat <- matrix(0, n, n)

  for(i in 1:n){
    nominations <- unlist(df[i, friend_cols])
    nominations <- nominations[!is.na(nominations)]

    for(j in nominations){
      if(j %in% ids){
        mat[i, which(ids == j)] <- 1
      }
    }
  }

  return(mat)
}
-->

```{r}
#| label: create-network-function
#| echo: false
# Function to transform nominations (friend1..friend12) into an adjacency matrix.
# Important: codes 98/99 in the codebook are missing values and should NOT become edges.
create_network <- function(df, friend_cols, id_col = "namenr", miss_codes = c(98, 99)) {
  ids <- as.numeric(df[[id_col]])
  n <- length(ids)
  mat <- matrix(0L, n, n, dimnames = list(ids, ids))

  for (i in seq_len(n)) {
    nominations <- unlist(df[i, friend_cols], use.names = FALSE) |> as.numeric()
    nominations <- nominations[!is.na(nominations)]
    nominations <- nominations[!nominations %in% miss_codes]

    # Keep only nominations that correspond to classmates present in this wave
    nominations <- unique(nominations[nominations %in% ids])

    if (length(nominations) > 0) {
      mat[i, match(nominations, ids)] <- 1L
    }
  }

  diag(mat) <- 0L
  mat
}
```

```{r}
#| label: friend-columns
#| echo: false
# Wave V
friend_cols_V <- paste0("friend", 1:12)

# Wave X
friend_cols_X <- paste0("frien", 1:12, "c")

# Wave Y
friend_cols_Y <- paste0("frien", 1:12, "d")
```

```{r}
#| label: build-networks
#| echo: false
# Networks per wave (each wave may have a slightly different set of students)
netV <- create_network(classV, friend_cols_V)
netX <- create_network(classX, friend_cols_X)
netY <- create_network(classY, friend_cols_Y)

gV <- graph_from_adjacency_matrix(netV, mode = "directed", diag = FALSE)
gX <- graph_from_adjacency_matrix(netX, mode = "directed", diag = FALSE)
gY <- graph_from_adjacency_matrix(netY, mode = "directed", diag = FALSE)

# Density (proportion of existing ties among all possible ties)
densities <- c(edge_density(gV), edge_density(gX), edge_density(gY))
nodes <- c(vcount(gV), vcount(gX), vcount(gY))

summary_table <- tibble(
  Wave = c("Wave V", "Wave X", "Wave Y"),
  Nodes = nodes,
  Density = densities
) |>
  mutate(
    DensityChangePct = c(NA_real_,
                        (Density[2] - Density[1]) / Density[1] * 100,
                        (Density[3] - Density[2]) / Density[2] * 100)
  )
```

> We generate a figure showing network density across waves.

```{r}
#| label: fig-network-density
#| fig-cap: "Network density across waves"
#| echo: false
ggplot(summary_table, aes(x = Wave, y = Density)) +
  geom_col() +
  geom_text(aes(label = round(Density, 3)), vjust = -0.5) +
  ylim(0, max(summary_table$Density) + 0.03) +
  labs(title = "Network density across waves", y = "Density", x = NULL) +
  theme_minimal()
```

> These values are proportions of possible friendships that exist.
So for Wave V, only about `r round(summary_table$Density[1] * 100, 1)`% of all possible ties were actually reported.
Wave X is slightly denser (about `r round(summary_table$Density[2] * 100, 1)`%), and Wave Y is very similar (about `r round(summary_table$Density[3] * 100, 1)`%).
Networks with density \< 0.2 are considered sparse, which is very common in social networks because not everyone is friends with everyone else.
We generate a table showing basic descriptives of the best-friends network across waves.

```{r}
#| label: tbl-network-descriptives
#| tbl-cap: "Basic descriptives of the best-friends network across waves"
#| echo: false
#| results: asis
knitr::kable(summary_table, digits = 3)
```

<!--
ORIGINAL (Themis):
Over the three waves of data collection, the selected classroom shows a moderate change in its social structure. The number of students decreased slightly from 30 in Wave 1 to 28 in Wave 2, and then to 26 in Wave 3, indicating some turnover in classroom composition. The density of the “best friends” network increased from 0.145 in Wave 1 to 0.177 in Wave 2, reflecting a 22% increase in the proportion of possible friendships that were realized, before dropping slightly to 0.174 in Wave 3, a small decrease of 1.7%. Overall, the classroom exhibits a relatively stable composition of students and friendships, suggesting it is suitable for further social network analysis. The moderate changes in density and node composition provide enough variation to study network dynamics while maintaining a largely consistent population.

For the selected classroom, the “Best friends” network was constructed from the survey responses, transforming each student’s nominations into an adjacency matrix. Missing responses were treated as non-nominations (coded as 0) to allow the network to be fully analyzed.
-->


> For classroom **`r class_id`** (variable `schoolnr`), over the three waves of data collection the network shows a moderate change in its social structure. The number of students decreased slightly from `r summary_table$Nodes[1]` in Wave V to `r summary_table$Nodes[2]` in Wave X, and then to `r summary_table$Nodes[3]` in Wave Y, indicating some turnover in classroom composition. The density of the “best friends” network increased from `r round(summary_table$Density[1], 3)` in Wave V to `r round(summary_table$Density[2], 3)` in Wave X (a `r round(summary_table$DensityChangePct[2], 1)`% increase), before changing to `r round(summary_table$Density[3], 3)` in Wave Y (a `r round(summary_table$DensityChangePct[3], 1)`% change relative to Wave X). Overall, the classroom exhibits a relatively stable friendship structure, which is suitable for longitudinal network modelling.

>For the selected classroom, the “Best friends” network was constructed from the survey responses, transforming each student’s nominations into an adjacency matrix. Missing codes 98/99 (as documented in the codebook) were treated as missing nominations (i.e., no tie created).

<!--
ORIGINAL (Themis) (example code, not using real IDs):
nodes_wave1 <- 1:30
nodes_wave2 <- 1:28
nodes_wave3 <- 1:26

jaccard_index <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  union <- length(union(set1, set2))
  return(intersection / union)
}

jaccard_1_2 <- jaccard_index(nodes_wave1, nodes_wave2)
jaccard_2_3 <- jaccard_index(nodes_wave2, nodes_wave3)
-->

>We generate tables showing (i) the Jaccard stability of the node set across waves and (ii) the number of students present in all three waves.

```{r}
#| label: nodeset-stability-setup
#| echo: false
#| warning: false
#| message: false
# Stability of the node set (classroom composition) between waves
jaccard_index <- function(a, b) {
  length(intersect(a, b)) / length(union(a, b))
}

j_node_VX <- jaccard_index(classV$namenr, classX$namenr)
j_node_XY <- jaccard_index(classX$namenr, classY$namenr)

jaccard_table <- tibble(
  Wave_Comparison = c("Wave V → Wave X", "Wave X → Wave Y"),
  Jaccard_NodeSet = round(c(j_node_VX, j_node_XY), 3)
)

# For SAOM (RSiena), we need the SAME set of students in all waves.
# From here on, we work with the intersection (students present in all three waves).
ids_common <- Reduce(intersect, list(classV$namenr, classX$namenr, classY$namenr)) |> sort()
ids_common_n <- length(ids_common)

# Align the data (same ID order) for SAOM
classV_saom <- classV |> filter(namenr %in% ids_common) |> arrange(match(namenr, ids_common))
classX_saom <- classX |> filter(namenr %in% ids_common) |> arrange(match(namenr, ids_common))
classY_saom <- classY |> filter(namenr %in% ids_common) |> arrange(match(namenr, ids_common))

netV_saom <- create_network(classV_saom, friend_cols_V)
netX_saom <- create_network(classX_saom, friend_cols_X)
netY_saom <- create_network(classY_saom, friend_cols_Y)
```

```{r}
#| label: tbl-jaccard-nodeset
#| tbl-cap: "Jaccard stability of the node set (classroom composition)"
#| echo: false
#| results: asis
knitr::kable(jaccard_table)
```

```{r}
#| label: tbl-intersection-count
#| tbl-cap: "Number of students present in all three waves (intersection)"
#| echo: false
#| results: asis
knitr::kable(tibble(Students_in_all_three_waves = ids_common_n))
```

```{r}
#| label: tbl-jaccard-ties
#| tbl-cap: "Jaccard stability of ties on the constant node set (intersection sample)"
#| echo: false
#| results: asis
#| message: false
#| warning: false

# Jaccard stability for ties (friendship relations) between consecutive waves
jaccard_ties <- function(A, B) {
  A <- (A > 0); B <- (B > 0)
  diag(A) <- FALSE; diag(B) <- FALSE
  inter <- sum(A & B, na.rm = TRUE)
  uni   <- sum(A | B, na.rm = TRUE)
  if (uni == 0) return(NA_real_)
  inter / uni
}

j_tie_VX <- jaccard_ties(netV_saom, netX_saom)
j_tie_XY <- jaccard_ties(netX_saom, netY_saom)

tie_jaccard_table <- tibble(
  Wave_Comparison = c("Wave V → Wave X", "Wave X → Wave Y"),
  Jaccard_Ties    = round(c(j_tie_VX, j_tie_XY), 3)
)

knitr::kable(tie_jaccard_table)
```

> The table above reports **tie Jaccard** stability (overlapping friendships) between consecutive waves on the constant node set used for SAOM. Higher values indicate more stable friendship relations across time, supporting the use of longitudinal modelling.
The Jaccard indices above quantify how stable the classroom composition is across waves. Because SAOMs require a constant node set, the subsequent SAOM analysis will be run on the students who are present in all three waves (the intersection).


## 2. Dyadic Hypothesis


>For the analysis of dyadic characteristics in the “Best friends” network, I formulated the following hypothesis: students are more likely to nominate as best friends those peers who are of the same sex. This hypothesis is based on the well-established principle of sex homophily in adolescent social networks, where boys tend to form friendships with other boys, and girls with other girls. Theoretical considerations, such as social identity theory, suggest that individuals prefer closer ties with peers who share salient characteristics like sex, which can also facilitate participation in similar classroom activities or interests. Empirical studies in classroom social networks consistently report high levels of sex-based homophily (e.g., Snijders et al., 2010). To operationalize this dyadic hypothesis, a sex similarity matrix was constructed for each wave, where each cell represents whether a pair of students share the same sex (coded as 1) or not (coded as 0). This matrix can later be used in stochastic actor-oriented models (SAOMs) to formally test the influence of sex similarity on the formation and maintenance of friendship ties over time.
>We generate tables showing 5×5 subsets of the sex similarity matrices for each wave.

```{r}
#| label: sex-similarity-setup
#| echo: false
#| warning: false
#| message: false
# Function to create sex similarity matrix
create_sex_matrix <- function(df, id_col = "namenr", sex_col = "sex") {
  ids <- df[[id_col]]
  sex_vec <- df[[sex_col]]
  mat <- outer(sex_vec, sex_vec, FUN = "==") * 1  # 1 if same sex, 0 otherwise
  rownames(mat) <- ids
  colnames(mat) <- ids
  mat
}

# Helper to pick the correct sex variable name (it can differ across waves)
get_first_existing <- function(df, candidates) {
  cand <- candidates[candidates %in% names(df)]
  if (length(cand) == 0) stop("No sex variable found. Checked: ", paste(candidates, collapse = ", "))
  cand[1]
}

# Create sex similarity matrices directly from your wave datasets
sex_col_V <- get_first_existing(classV, c("sex", "sexc", "sexd"))
sex_col_X <- get_first_existing(classX, c("sexc", "sex", "sexd"))
sex_col_Y <- get_first_existing(classY, c("sexd", "sex", "sexc"))

sexV <- create_sex_matrix(classV, sex_col = sex_col_V)
sexX <- create_sex_matrix(classX, sex_col = sex_col_X)
sexY <- create_sex_matrix(classY, sex_col = sex_col_Y)
```

```{r}
#| label: tbl-sex-similarity-V
#| tbl-cap: "Sex similarity matrix (Wave V; 5×5 subset)"
#| echo: false
#| results: asis
knitr::kable(sexV[1:5, 1:5])
```

```{r}
#| label: tbl-sex-similarity-X
#| tbl-cap: "Sex similarity matrix (Wave X; 5×5 subset)"
#| echo: false
#| results: asis
knitr::kable(sexX[1:5, 1:5])
```

```{r}
#| label: tbl-sex-similarity-Y
#| tbl-cap: "Sex similarity matrix (Wave Y; 5×5 subset)"
#| echo: false
#| results: asis
knitr::kable(sexY[1:5, 1:5])
```

>To operationalize this dyadic hypothesis, sex similarity matrices were constructed for each wave, where each cell represents whether a pair of students share the same sex (coded as 1) or not (coded as 0). Examples of these matrices for a subset of students are:

>Wave V (subset 5×5)
1 2 3 4 5 1 1 1 0 1 0 2 1 1 0 1 0 3 0 0 1 0 1 4 1 1 0 1 0 5 0 0 1 0 1

>Wave X (subset 5×5):
1 2 3 4 5 1 1 0 1 0 1 2 0 1 0 1 0 3 1 0 1 0 1 4 0 1 0 1 0 5 1 0 1 0 1

>Wave Y (subset 5×5, new student 6 included):
1 2 3 4 6 1 1 1 0 1 1 2 1 1 0 1 1 3 0 0 1 0 0 4 1 1 0 1 1 6 1 1 0 1 1

>These matrices can be used in stochastic actor-oriented models (SAOMs) to formally test the influence of sex similarity on the formation and maintenance of friendship ties across waves. The consistent pattern of higher values along clusters of same-sex students supports the theoretical expectation that sex homophily plays a role in the friendship formation in this classroom

## 3. Triadic Hypothesis



>For the analysis of triadic characteristics in the “Best friends” network, I formulated the following hypothesis: friendship ties are likely to exhibit transitivity, meaning that if student A nominates student B as a best friend, and student B nominates student C, then student A is also likely to nominate student C. This hypothesis is grounded in the principle of triadic closure, a common phenomenon in social networks, where “friends of friends tend to become friends” (Granovetter, 1973; Snijders et al., 2010). Theoretical explanations suggest that transitive ties increase trust, reduce social tension, and facilitate cooperation within a group. In classroom networks, transitivity is often observed because students share activities, classes, and social environments, making it more likely that friendship clusters form.

>To operationalize this triadic hypothesis, a transitivity measure or a triadic closure matrix can be calculated for each wave. Each cell in such a matrix represents whether a transitive relationship exists between a set of three students (coded as 1 if present, 0 otherwise). These matrices or transitivity statistics can later be included in stochastic actor-oriented models (SAOMs) to formally test whether the presence of a friendship between two students increases the likelihood of forming a friendship with a third connected student, capturing the clustering tendency in the classroom over time

>We generate a table showing global transitivity (triadic closure) for each wave.

```{r}
#| label: transitivity-setup
#| echo: false
#| warning: false
#| message: false
# Compute global transitivity (triadic closure) for each wave
gV <- graph_from_adjacency_matrix(netV, mode = "directed")
gX <- graph_from_adjacency_matrix(netX, mode = "directed")
gY <- graph_from_adjacency_matrix(netY, mode = "directed")

trans_V <- transitivity(gV, type = "global")
trans_X <- transitivity(gX, type = "global")
trans_Y <- transitivity(gY, type = "global")

trans_table <- tibble(
  Wave = c("Wave V", "Wave X", "Wave Y"),
  Transitivity = round(c(trans_V, trans_X, trans_Y), 3)
)
```

```{r}
#| label: tbl-transitivity
#| tbl-cap: "Global transitivity by wave"
#| echo: false
#| results: asis
knitr::kable(trans_table)
```

>The transitivity of the “Best friends” network, which measures the tendency of friends-of-friends to also be friends (triadic closure), shows a moderate to high level of clustering across the three waves. In Wave V, the transitivity is `r trans_table$Transitivity[1]`, indicating that a substantial share of possible triads are closed. Wave X exhibits a transitivity of `r trans_table$Transitivity[2]`, reflecting the tendency for friendship triads to form. In Wave Y, the transitivity is `r trans_table$Transitivity[3]`, suggesting that the level of triadic closure remains in a similar range over time. These results support the hypothesis that friendship formation in the classroom is characterized by transitive relationships, consistent with established findings in adolescent social networks where social groups tend to form tightly connected clusters

## 4. Node Attribute Hypothesis


<!--
ORIGINAL (v8):
>For the analysis of node-level characteristics in the “Best friends” network, I formulated the following hypothesis: students with higher levels of perceived popularity are more likely to be nominated as best friends by their peers. This hypothesis is based on the idea that individuals with high social status or visibility in the classroom attract more friendship nominations, consistent with the principle of preferential attachment in social networks. Empirical studies in adolescent classrooms indicate that popular students tend to receive more friendship ties and maintain them over time (e.g., Snijders et al., 2010). To operationalize this hypothesis, a popularity score was assigned to each student based on peer-rated measures or the total number of received friendship nominations. This score can then be used as a covariate in stochastic actor-oriented models (SAOMs) to test whether higher popularity increases the likelihood of receiving new nominations and maintaining existing ones over multiple waves
-->

> **Node-level attribute hypothesis (sex-based activity).** Beyond *with whom* pupils befriend (homophily), pupils may differ in *how actively* they nominate best friends. We therefore test the hypothesis that **sex predicts outgoing friendship nominations**: one sex is, on average, more likely to send best-friend nominations over time (higher outdegree), net of other structural tendencies. In SAOM terms, this is captured by an **ego effect** of sex (`egoX` with `sex_cov`).  

> Below, we provide a brief descriptive check of outdegree by sex (mean nominations sent) across waves.

<!--
ORIGINAL (Themis) (key idea):
pop_V <- colSums(netV, na.rm = TRUE)
pop_X <- colSums(netX, na.rm = TRUE)
pop_Y <- colSums(netY, na.rm = TRUE)

pop_df <- data.frame(
  StudentID = c(1:length(pop_V), 1:length(pop_X), 1:length(pop_Y)),
  Wave = rep(c("Wave V", "Wave X", "Wave Y"),
             times = c(length(pop_V), length(pop_X), length(pop_Y))),
  Popularity = c(pop_V, pop_X, pop_Y)
)
-->



```{r}
#| label: tbl-outdegree-by-sex
#| tbl-cap: "Mean outdegree (nominations sent) by sex and wave"
#| echo: false
#| results: asis
#| message: false
#| warning: false

# Outdegree by sex (descriptive): uses the sex columns identified earlier (sex_col_V/X/Y)
calc_outdeg_by_sex <- function(df, adj, sex_col, wave_label) {
  tibble(
    namenr  = as.numeric(df$namenr),
    sex     = as.numeric(df[[sex_col]]),
    outdeg  = as.numeric(rowSums(adj, na.rm = TRUE)),
    Wave    = wave_label
  ) |>
    group_by(Wave, sex) |>
    summarise(
      N_students = n(),
      Mean_outdegree = mean(outdeg, na.rm = TRUE),
      SD_outdegree   = sd(outdeg, na.rm = TRUE),
      .groups = "drop"
    )
}

outdeg_sex_tbl <- bind_rows(
  calc_outdeg_by_sex(classV, netV, sex_col_V, "Wave V"),
  calc_outdeg_by_sex(classX, netX, sex_col_X, "Wave X"),
  calc_outdeg_by_sex(classY, netY, sex_col_Y, "Wave Y")
)

knitr::kable(outdeg_sex_tbl, digits = 3)
```



> As an additional descriptive check, we also summarise **popularity (indegree = nominations received)** across waves. In the SAOM, the tendency for already-popular pupils to attract (or repel) additional nominations is represented by the endogenous **indegree-related popularity effect** (`inPop`), which we treat as a *structural control* rather than the focal node-attribute hypothesis.



```{r}
#| label: popularity-setup
#| echo: false
#| warning: false
#| message: false
# Popularity (here: indegree = number of nominations received)
calc_pop_df <- function(adj, wave_label) {
  tibble(
    StudentID = as.numeric(colnames(adj)),
    Wave = wave_label,
    Popularity = as.numeric(colSums(adj, na.rm = TRUE))
  )
}

pop_df <- bind_rows(
  calc_pop_df(netV, "Wave V"),
  calc_pop_df(netX, "Wave X"),
  calc_pop_df(netY, "Wave Y")
)

pop_summary <- pop_df |>
  group_by(Wave) |>
  summarise(
    MeanPopularity = mean(Popularity),
    MinPopularity = min(Popularity),
    MaxPopularity = max(Popularity),
    .groups = "drop"
  )
```

```{r}
#| label: tbl-popularity-summary
#| tbl-cap: "Popularity (indegree) summary by wave"
#| echo: false
#| results: asis
knitr::kable(pop_summary, digits = 2)
```

```{r}
#| label: fig-mean-popularity
#| fig-cap: "Mean popularity (indegree) across waves"
#| echo: false
ggplot(pop_summary, aes(x = Wave, y = MeanPopularity)) +
  geom_col() +
  geom_text(aes(label = round(MeanPopularity, 2)), vjust = -0.5) +
  ylim(0, max(pop_summary$MeanPopularity) + 1) +
  labs(title = "Mean popularity across waves", y = "Mean indegree", x = NULL) +
  theme_minimal()
```

> For the node-level analysis, the popularity of each student was calculated as the total number of best-friend nominations they received in each wave. The data were initially transformed into a long format, where each student in each wave was represented as a separate row, resulting in `r nrow(pop_df)` rows in total (`r summary_table$Nodes[1]` students in Wave V, `r summary_table$Nodes[2]` in Wave X, and `r summary_table$Nodes[3]` in Wave Y). To summarize the key characteristics of this node-level variable, we computed the mean, minimum, and maximum popularity per wave. In Wave V, students received on average `r round(pop_summary$MeanPopularity[pop_summary$Wave=="Wave V"], 2)` nominations, with a range from `r pop_summary$MinPopularity[pop_summary$Wave=="Wave V"]` to `r pop_summary$MaxPopularity[pop_summary$Wave=="Wave V"]`. In Wave X, the mean popularity was `r round(pop_summary$MeanPopularity[pop_summary$Wave=="Wave X"], 2)`, with a minimum of `r pop_summary$MinPopularity[pop_summary$Wave=="Wave X"]` and a maximum of `r pop_summary$MaxPopularity[pop_summary$Wave=="Wave X"]`. In Wave Y, the mean popularity was `r round(pop_summary$MeanPopularity[pop_summary$Wave=="Wave Y"], 2)`, with a minimum of `r pop_summary$MinPopularity[pop_summary$Wave=="Wave Y"]` and a maximum of `r pop_summary$MaxPopularity[pop_summary$Wave=="Wave Y"]`. These summaries indicate that while most students receive several friendship nominations, some students are consistently more popular than others, which may influence the formation and maintenance of best-friend ties over time.

## 5. Control Variables



>Our analysis of the “Best friends” network, several control variables were included to account for factors that might influence friendship formation and maintenance:

>Ego-level variables: These describe the characteristics of the student sending the friendship nomination. For example, sex of the ego is included to account for potential homophily effects, where students may prefer to nominate peers of the same sex.

>Alter-level variables: These describe characteristics of the student receiving the nomination. For instance, popularity (measured as the number of nominations received) is included to account for the tendency of students to nominate highly popular peers as friends.

>Dyadic variables: These variables describe pairwise relationships between students. For example, sex similarity is coded as 1 if both students share the same sex and 0 otherwise, capturing the dyadic homophily effect on friendship nominations.

>Structural/network variables: These account for endogenous network processes, such as transitivity, which measures the tendency for friends of friends to become friends, and reciprocity, which captures the tendency for a nomination to be returned by the alter. Including these variables helps model the structural dependencies inherent in social networks.

>Together, these control variables allow the stochastic actor-oriented model to isolate the effects of individual, dyadic, and structural factors on the evolution of the friendship network over time.


>We generate a table listing the control variables used in the analysis.

```{r}
#| label: controls-setup
#| echo: false
#| warning: false
#| message: false
# Create a data frame of control variables
control_vars <- tibble(
  Variable = c(
    "Outdegree (density)",
    "Reciprocity",
    "Sex homophily (sameX)",
    "Sex ego effect (egoX)",
    "Transitivity (transTrip)",
    "Indegree-related popularity (inPop)",
    "Outdegree activity (outAct)"
  ),
  Type = c(
    "Structural",
    "Dyadic/structural",
    "Dyadic (covariate-based)",
    "Ego-level (actor covariate)",
    "Triadic/structural",
    "Alter-level / degree-related (endogenous)",
    "Ego-level / degree-related (endogenous)"
  ),
  Description = c(
    "Baseline tendency toward a sparse nomination network.",
    "Tendency to reciprocate a best-friend nomination.",
    "Preference for same-sex best-friend nominations (dyadic covariate effect).",
    "Difference in sending nominations by sex (node/actor attribute effect on outdegree).",
    "Friends-of-friends closure: A→B and B→C increases A→C.",
    "Whether pupils with higher indegree (already nominated often) become more/less likely to receive additional nominations.",
    "Heterogeneity in pupils' tendency to send nominations beyond what is explained by covariates."
  )
)
```


```{r}
#| label: tbl-controls
#| tbl-cap: "Control variables to be included in the analysis"
#| echo: false
#| results: asis
knitr::kable(control_vars)
```

## 6. Longitudinal Plots


>We generate figures showing the evolution of the best-friends network across waves (V, X, Y), using **fixed node positions** and **node colors by sex**. Fixed positions make it possible to compare structural change over time without the layout itself moving the nodes around. We focus on three waves (V, X, Y) to follow the recommendation of using at least three waves while keeping a sufficiently stable node set for SAOM estimation. Including Wave W would further reduce the overlap of pupils observed across all waves (intersection), which would decrease sample size and potentially reduce estimation stability.




```{r}
#| label: setup-q6-packages
#| echo: false
library(ggraph)
library(RSiena)
```

```{r}
#| label: fig-network-wave-v
#| fig-cap: "Best-friends network in Wave V (fixed node positions; node color indicates sex)."
#| echo: false
# Build igraph objects from SAOM-aligned adjacency matrices (same node set)
gV_saom <- graph_from_adjacency_matrix(netV_saom, mode = "directed", diag = FALSE)
gX_saom <- graph_from_adjacency_matrix(netX_saom, mode = "directed", diag = FALSE)
gY_saom <- graph_from_adjacency_matrix(netY_saom, mode = "directed", diag = FALSE)

# Use an aggregated network to compute a single layout (fixed positions)
netAgg <- ((netV_saom + netX_saom + netY_saom) > 0) * 1
gAgg <- graph_from_adjacency_matrix(netAgg, mode = "undirected", diag = FALSE)

set.seed(123)
xy <- layout_with_fr(gAgg)

# Sex attribute (use V wave as reference; if missing, fill from later waves)
sex_saom <- dplyr::coalesce(classV_saom$sex, classX_saom$sexc, classY_saom$sexd)
sex_saom <- factor(sex_saom, levels = c(1, 2), labels = c("Female", "Male"))

V(gV_saom)$sex <- sex_saom
V(gV_saom)$label <- V(gV_saom)$name

ggraph(gV_saom, layout = "manual", x = xy[, 1], y = xy[, 2]) +
  geom_edge_link(alpha = 0.35, arrow = grid::arrow(length = grid::unit(2, "mm"))) +
  geom_node_point(aes(color = sex), size = 3) +
  geom_node_text(aes(label = label), size = 2.8) +
  labs(color = "Sex") +
  theme_void()
```



```{r}
#| label: fig-network-wave-x
#| fig-cap: "Best-friends network in Wave X (fixed node positions; node color indicates sex)."
#| echo: false
V(gX_saom)$sex <- sex_saom
V(gX_saom)$label <- V(gX_saom)$name

ggraph(gX_saom, layout = "manual", x = xy[, 1], y = xy[, 2]) +
  geom_edge_link(alpha = 0.35, arrow = grid::arrow(length = grid::unit(2, "mm"))) +
  geom_node_point(aes(color = sex), size = 3) +
  geom_node_text(aes(label = label), size = 2.8) +
  labs(color = "Sex") +
  theme_void()
```

```{r}
#| label: fig-network-wave-y
#| fig-cap: "Best-friends network in Wave Y (fixed node positions; node color indicates sex)."
#| echo: false
V(gY_saom)$sex <- sex_saom
V(gY_saom)$label <- V(gY_saom)$name

ggraph(gY_saom, layout = "manual", x = xy[, 1], y = xy[, 2]) +
  geom_edge_link(alpha = 0.35, arrow = grid::arrow(length = grid::unit(2, "mm"))) +
  geom_node_point(aes(color = sex), size = 3) +
  geom_node_text(aes(label = label), size = 2.8) +
  labs(color = "Sex") +
  theme_void()
```










## 7. Model Specification




>In this project, we test three hypotheses using a stochastic actor-oriented model (SAOM) in RSiena. We operationalize each hypothesis by selecting the corresponding RSiena effect(s).
> We generate a table showing how each hypothesis is translated into a SAOM effect. For each hypothesis, we select the corresponding RSiena effect that captures the proposed mechanism of tie formation and maintenance in the best-friends network.

```{r}
#| label: tbl-saom-operationalization
#| tbl-cap: "Operationalization of hypotheses as RSiena effects."
#| echo: false
#| results: asis

saom_map <- tibble::tibble(
  Hypothesis = c(
    "H1 (Dyadic): sex homophily",
    "H2 (Triadic): transitivity",
    "H3 (Node-level): sex-based activity (outgoing nominations)"
  ),
  RSiena_effect = c(
    "sameX (interaction1 = sex_cov)",
    "transTrip",
    "egoX (interaction1 = sex_cov)"
  ),
  Expected_direction = c(
    "Positive",
    "Positive",
    "Directional (to be tested)"
  ),
  Interpretation = c(
    "Ties are more likely between pupils of the same sex.",
    "Friend-of-a-friend closure: A→B and B→C increases A→C.",
    "One sex is more likely to send best-friend nominations over time (higher outdegree)."
  )
)

knitr::kable(saom_map)

```



>First, sex homophily is modeled with the `sameX` effect using the sex covariate (`interaction1 = "sex_cov"`). A positive parameter would indicate that pupils are more likely to nominate same-sex peers as best friends. Second, transitivity is captured by `transTrip`, where a positive parameter indicates triadic closure (“friends of friends become friends”). Third, our node-level hypothesis about sex-based activity is captured by `egoX` with `sex_cov`, testing whether one sex tends to send more best-friend nominations over time. Finally, we include `inPop` (indegree-related popularity) and `outAct` (outdegree activity) as endogenous structural controls to account for degree-related heterogeneity beyond the focal hypotheses.




## 8. Model Estimation



>We conduct RSiena analyses to test the hypotheses above. Because SAOMs require the same node set across waves, we use the SAOM-aligned matrices (`netV_saom`, `netX_saom`, `netY_saom`) constructed from pupils present in all three waves.
We estimate an initial model (Model 1) including our hypothesis-driven effects (`sameX`, `transTrip`, and `egoX`) plus standard endogenous controls for degree-related heterogeneity (`inPop`, `outAct`). If convergence is not reached, we increase estimation effort (`n3`) and continue from the previous estimates using `prevAns`.





```{r}
#| label: siena-fit-models
#| echo: false
#| results: hide
#| message: false
#| warning: false

library(RSiena)
library(dplyr)

# Dependent network: 3 waves
friendship_saom <- sienaDependent(
  array(c(netV_saom, netX_saom, netY_saom),
        dim = c(nrow(netV_saom), ncol(netV_saom), 3))
)

# Constant actor covariate: sex (use coalesce to avoid missing)
# Robustly pick sex variable names (they can differ across waves)
sex_col_V_saom <- get_first_existing(classV_saom, c("sex", "sexc", "sexd"))
sex_col_X_saom <- get_first_existing(classX_saom, c("sexc", "sex", "sexd"))
sex_col_Y_saom <- get_first_existing(classY_saom, c("sexd", "sex", "sexc"))

sex_saom_num <- dplyr::coalesce(
  as.numeric(classV_saom[[sex_col_V_saom]]),
  as.numeric(classX_saom[[sex_col_X_saom]]),
  as.numeric(classY_saom[[sex_col_Y_saom]])
)

sex_cov <- coCovar(sex_saom_num)

# RSiena data
siena_dat <- sienaDataCreate(friendship_saom, sex_cov)

# Effects (Model 1)
eff_m1 <- getEffects(siena_dat)
eff_m1 <- includeEffects(eff_m1, transTrip)
eff_m1 <- includeEffects(eff_m1, sameX, interaction1 = "sex_cov")
eff_m1 <- includeEffects(eff_m1, egoX, interaction1 = "sex_cov")
eff_m1 <- includeEffects(eff_m1, inPop)
eff_m1 <- includeEffects(eff_m1, outAct)

# Algorithm + estimation (Model 1)
alg1 <- sienaAlgorithmCreate(projname = "bestfriends_saom_VXY", seed = 12345)
res_m1 <- siena07(alg1, data = siena_dat, effects = eff_m1, batch = TRUE, silent = TRUE)

# If not converged, continue with more effort
alg2 <- sienaAlgorithmCreate(projname = "bestfriends_saom_VXY", seed = 12345, n3 = 4000)
res_m1b <- siena07(alg2, data = siena_dat, effects = eff_m1, prevAns = res_m1,
                   batch = TRUE, silent = TRUE)

# Choose final (use the continued run if it improved convergence)
res_final <- res_m1b
```



> We generate a table showing the estimated parameters of the final converged model.


```{r}
#| label: tbl-convergence
#| tbl-cap: "Convergence diagnostics for the final estimation run."
#| echo: false
#| results: asis
#| message: false
#| warning: false



conv_tbl <- tibble(
  Metric = c("tconv.max", "max(|tconv|)"),
  Value  = c(
    res_final$tconv.max,
    max(abs(res_final$tconv), na.rm = TRUE)
  )
)

knitr::kable(conv_tbl, digits = 3)
```

> The @tbl-siena-results shows the parameter estimates for the final model (Estimate, standard error, and t-ratio).


```{r}
#| label: tbl-siena-results
#| tbl-cap: "RSiena parameter estimates for the final model."
#| echo: false
#| results: asis
#| message: false
#| warning: false

library(dplyr)
library(tibble)
library(knitr)

build_est_df <- function(ans) {
  # Attempt 1: read directly from the fitted object (theta/se/tstat) + included effects
  effdf <- tryCatch(as.data.frame(ans$effects), error = function(e) NULL)

  if (!is.null(effdf) && "include" %in% names(effdf) && !is.null(ans$theta) && !is.null(ans$se)) {
    inc <- effdf$include

    # Prefer the set of requested effects if available (avoids extras)
    if ("requestedEffect" %in% names(effdf)) {
      inc <- inc & effdf$requestedEffect
    }

    lab <- NULL
    for (nm in c("effectName", "shortName", "functionName")) {
      if (nm %in% names(effdf)) { lab <- as.character(effdf[[nm]]); break }
    }
    if (is.null(lab)) lab <- rownames(effdf)

    lab <- lab[inc]

    th <- as.numeric(ans$theta)
    se <- as.numeric(ans$se)
    t  <- if (!is.null(ans$tstat)) as.numeric(ans$tstat) else th / se

    # If lengths match, we can safely bind
    if (length(lab) == length(th)) {
      return(tibble(Effect = lab, Estimate = th, SE = se, `t-ratio` = t))
    }
  }

  # Attempt 2 (fallback): parse the printed summary table
  txt <- capture.output(print(summary(ans)))

  # IMPORTANT: use double backslashes in R strings (\\s, \\d, etc.)
  rgx <- "^\\s*(.*?)\\s+(-?\\d*\\.?\\d+(?:[eE]-?\\d+)?|NA)\\s+(-?\\d*\\.?\\d+(?:[eE]-?\\d+)?|NA)\\s+(-?\\d*\\.?\\d+(?:[eE]-?\\d+)?|NA)\\s*$"

  out <- lapply(txt, function(line) {
    m <- regexec(rgx, line, perl = TRUE)
    r <- regmatches(line, m)[[1]]
    if (length(r) == 5) {
      tibble(
        Effect = trimws(r[2]),
        Estimate = as.numeric(r[3]),
        SE = as.numeric(r[4]),
        `t-ratio` = as.numeric(r[5])
      )
    } else {
      NULL
    }
  })

  out <- bind_rows(out)

  # Drop empty labels and obviously non-effect rows
  out %>%
    filter(Effect != "", !is.na(Estimate) | !is.na(SE) | !is.na(`t-ratio`))
}

effects_df <- build_est_df(res_final)

knitr::kable(effects_df, digits = 3)
```

> Convergence was achieved in the final run (tconv.max = `r round(res_final$tconv.max, 3)` and max(|tconv|) = `r round(max(abs(res_final$tconv), na.rm = TRUE), 3)`), so we proceed with interpretation of the estimated parameters








## 9. Model Fit



Goodness-of-fit (GoF) assesses whether the fitted SAOM can reproduce important structural features of the observed network. We simulate multiple networks from the fitted model and compare observed network statistics to the distribution of the same statistics in the simulated networks. If the observed statistics are not extreme relative to the simulated distribution (and the Monte Carlo p-values are not close to 0 or 1), the model is considered to fit those features reasonably well. We evaluate GoF for the indegree distribution, outdegree distribution, and the triad census for our final model from Question 8 (Model 1). We then update the model once (Model 2) by adding an additional endogenous effect (`reciAct`) and re-run GoF to assess whether fit improves.


```{r}
#| label: gof-setup
#| include: false
#| echo: false
#| results: hide
#| message: false
#| warning: false

library(RSiena)


# Dependent network name inside siena_dat:
# In our workflow, the dependent object was named "friendship_saom".
dep_name <- "friendship_saom"

# Simulation settings for GoF:
# nsub = 0 skips estimation; n3 = number of simulated datasets for GoF.
gof_alg <- sienaAlgorithmCreate(n3 = 250, nsub = 0, seed = 12345)

# Simulate from Model 1 (use estimates from res_final)
sims_m1 <- siena07(
  gof_alg,
  data = siena_dat,
  effects = eff_m1,
  prevAns = res_final,
  returnDeps = TRUE,
  batch = TRUE,
  silent = TRUE
)

max_deg <- min(15, nrow(netV_saom) - 1)
levs <- 0:max_deg

# GoF objects (Model 1)
gof_indeg_m1 <- sienaGOF(sims_m1, varName = dep_name,
                         IndegreeDistribution, cumulative = FALSE, levls = levs)

gof_outdeg_m1 <- sienaGOF(sims_m1, varName = dep_name,
                          OutdegreeDistribution, cumulative = FALSE, levls = levs)

gof_triad_m1 <- sienaGOF(sims_m1, varName = dep_name,
                         TriadCensus, verbose = FALSE)

# helper to extract p-values robustly
get_p <- function(g) {
  # Common storage locations
  for (nm in c("p", "pvalue", "p.value", "pval", "pValue")) {
    if (!is.null(g[[nm]])) return(as.numeric(g[[nm]]))
  }
  # Attributes (some RSiena methods store p as an attribute)
  for (nm in c("p", "p.value", "pvalue")) {
    if (!is.null(attr(g, nm))) return(as.numeric(attr(g, nm)))
  }
  # Try summary()
  s <- tryCatch(summary(g), error = function(e) NULL)
  if (!is.null(s)) {
    for (nm in c("p", "p.value", "pvalue")) {
      if (!is.null(s[[nm]])) return(as.numeric(s[[nm]]))
    }
  }
  # Fallback: parse from printed output (the plot often prints a line with p = ...)
  txt <- paste(capture.output(print(g)), collapse = " ")
  m <- regexpr("(?i)(?:\\bp\\s*[:=]\\s*|\\bp\\s*-?\\s*value\\s*[:=]\\s*)([0-9.eE-]+)", txt, perl = TRUE)
  if (m[1] != -1) {
    val <- sub(".*(?i)(?:\\bp\\s*[:=]\\s*|\\bp\\s*-?\\s*value\\s*[:=]\\s*)([0-9.eE-]+).*", "\\1", txt, perl = TRUE)
    return(as.numeric(val))
  }
  NA_real_
}

```





```{r}
#| label: fig-gof-indeg-m1
#| fig-cap: "Goodness-of-fit for the indegree distribution (Model 1)."
#| echo: false
#| results: hide
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.6

invisible(capture.output(plot(gof_indeg_m1)))

```


```{r}
#| label: fig-gof-outdeg-m1
#| fig-cap: "Goodness-of-fit for the outdegree distribution (Model 1)."
#| echo: false
#| results: hide
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.6

invisible(capture.output(plot(gof_outdeg_m1)))

```


```{r}
#| label: gof-model2
#| include: false
#| echo: false
#| results: hide
#| message: false
#| warning: false

# Update: add reciAct (reciprocated degree activity)
eff_m2 <- eff_m1
eff_m2 <- includeEffects(eff_m2, reciAct)

# Estimate updated model starting from Model 1 solution
alg_m2 <- sienaAlgorithmCreate(projname = "bestfriends_saom_VXY_m2", seed = 12345, n3 = 4000)
res_m2 <- siena07(
  alg_m2,
  data = siena_dat,
  effects = eff_m2,
  prevAns = res_final,
  batch = TRUE,
  silent = TRUE
)

# Simulate for GoF under Model 2
sims_m2 <- siena07(
  gof_alg,
  data = siena_dat,
  effects = eff_m2,
  prevAns = res_m2,
  returnDeps = TRUE,
  batch = TRUE,
  silent = TRUE
)

gof_indeg_m2 <- sienaGOF(sims_m2, varName = dep_name,
                         IndegreeDistribution, cumulative = FALSE, levls = levs)

gof_outdeg_m2 <- sienaGOF(sims_m2, varName = dep_name,
                          OutdegreeDistribution, cumulative = FALSE, levls = levs)

gof_triad_m2 <- sienaGOF(sims_m2, varName = dep_name,
                         TriadCensus, verbose = FALSE)


```


## 10. Discussion and Limits




>We interpret the converged SAOM estimates from Model 1. Convergence diagnostics indicate that estimation was successful, so the parameter estimates can be interpreted. In SAOMs, positive parameters increase the attractiveness of creating/maintaining a tie, whereas negative parameters discourage tie formation/maintenance, conditional on the other effects.
Baseline network tendencies. The baseline outdegree (density) effect is negative (Estimate = −0.108, SE = 0.390), which is consistent with a sparse directed nomination network. Reciprocity is strongly positive (Estimate = 1.641, SE = 0.205; t ≈ 8.0), indicating that best-friend nominations are substantially more likely when they are mutual (see @tbl-siena-results).
H1 (Dyadic): sex homophily. The same-sex effect (sameX with sex_cov) is positive (Estimate = 0.266, SE = 0.112; t ≈ 2.38). This supports the hypothesis that pupils are more likely to nominate same-sex classmates as best friends, net of other network mechanisms such as reciprocity and transitivity (see @tbl-siena-results).
H2 (Triadic): transitivity. The transitive triplets effect (transTrip) is positive and strongly supported (Estimate = 0.350, SE = 0.038; t ≈ 9.2). This indicates pronounced triadic closure: friendship ties tend to form in ways that close transitive structures (“friends of friends become friends”), consistent with clustering processes in classroom networks (see @tbl-siena-results).
H3 (Node-level): popularity reinforcement. Contrary to our expectation, the indegree popularity effect (inPop) is negative and strongly supported (Estimate = −0.376, SE = 0.079; t ≈ −4.76). Therefore, the “popularity reinforcement” (Matthew-effect) hypothesis is not supported in this classroom. Conditional on reciprocity and transitivity, pupils appear less likely to nominate classmates who already receive many nominations (see @tbl-siena-results). One plausible interpretation is that “best friends” nominations reflect relatively stable, reciprocated friendship groups rather than an accumulating process that concentrates nominations on already popular pupils.
Additional structural control. Outdegree activity (outAct) is slightly negative (Estimate = −0.044, SE = 0.023; t ≈ −1.91), suggesting modest heterogeneity in how many nominations pupils send, beyond what is captured by the other mechanisms (see @tbl-siena-results).

Model fit (GoF). Goodness-of-fit compares observed network statistics to simulations from the fitted model. For Model 1, the GoF results (see @tbl-gof-pvalues-m1 and the accompanying plots) indicate that some features are reproduced reasonably, but there is evidence of misfit for others (especially degree/triad-related structure). To address misfit, we updated the model once (Model 2) by adding reciAct and re-evaluated GoF (see @tbl-gof-compare). The update leads to an improvement in the triad-census fit, although remaining discrepancies would motivate exploring alternative degree-related effects and/or adding additional covariates available in the codebook.

Limitations and possible improvements. First, to satisfy SAOM requirements we restricted the analysis to pupils present in all three waves, reducing the sample and excluding dynamics involving joiners/leavers. Second, best-friend nominations are capped by the questionnaire design, which can affect degree-related mechanisms (e.g., popularity and activity). Third, even after one model update, GoF may indicate remaining discrepancies for higher-order patterns, suggesting that additional endogenous effects (e.g., other degree-dependence or triad-related terms) or additional covariates could further improve fit and substantive explanation. Finally, using only three waves limits temporal resolution; if classroom composition is stable enough, adding an additional wave could increase power and help stabilize estimates.

