---
title: "Final project "
author: " Themis Oluwatomole Ejegreh $ patricia  givort cruz cabral"
date: "2026-02-28"
format:
  pdf:
    number-sections: false
    fig-cap-location: top
    tbl-cap-location: top
    fig-numbering: true
    tbl-numbering: true
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
---

## Introduction 























## Descriptive Statistics

```{r}
#| label: setup-1
#| include: false
#| echo: false
#| warning: false
#| message: false
library(haven)
library(tidyverse)
library(igraph)
library(ggplot2)

# Adjust the path below to where your .sav files are stored.
# Tip: if the .sav files are in the same folder as this .qmd, use data_dir <- "."
data_dir <- "."

waveV_path <- file.path(data_dir, "PupilsWaveV.sav")
waveW_path <- file.path(data_dir, "PupilsWaveW_geanonimiseerd.sav")
waveX_path <- file.path(data_dir, "PupilsWaveX.sav")
waveY_path <- file.path(data_dir, "PupilsWaveY.sav")

waveV <- read_sav(waveV_path)
waveW <- read_sav(waveW_path)
waveX <- read_sav(waveX_path)
waveY <- read_sav(waveY_path)
```

```{r}
#| label: select-class
#| echo: false
class_id <- "09f"

classV <- subset(waveV, schoolnr == class_id)
classX <- subset(waveX, schoolnr == class_id)
classY <- subset(waveY, schoolnr == class_id)
```

completed

<!--
(Themis):
create_network <- function(df, friend_cols){

  ids <- df$namenr      # student IDs
  n <- length(ids)
  mat <- matrix(0, n, n)

  for(i in 1:n){
    nominations <- unlist(df[i, friend_cols])
    nominations <- nominations[!is.na(nominations)]

    for(j in nominations){
      if(j %in% ids){
        mat[i, which(ids == j)] <- 1
      }
    }
  }

  return(mat)
}
-->

```{r}
#| label: create-network-function
#| echo: false
# Function to transform nominations (friend1..friend12) into an adjacency matrix.
# Important: codes 98/99 in the codebook are missing values and should NOT become edges.
create_network <- function(df, friend_cols, id_col = "namenr", miss_codes = c(98, 99)) {
  ids <- as.numeric(df[[id_col]])
  n <- length(ids)
  mat <- matrix(0L, n, n, dimnames = list(ids, ids))

  for (i in seq_len(n)) {
    nominations <- unlist(df[i, friend_cols], use.names = FALSE) |> as.numeric()
    nominations <- nominations[!is.na(nominations)]
    nominations <- nominations[!nominations %in% miss_codes]

    # Keep only nominations that correspond to classmates present in this wave
    nominations <- unique(nominations[nominations %in% ids])

    if (length(nominations) > 0) {
      mat[i, match(nominations, ids)] <- 1L
    }
  }

  diag(mat) <- 0L
  mat
}
```

```{r}
#| label: friend-columns
#| echo: false
# Wave V
friend_cols_V <- paste0("friend", 1:12)

# Wave X
friend_cols_X <- paste0("frien", 1:12, "c")

# Wave Y
friend_cols_Y <- paste0("frien", 1:12, "d")
```

```{r}
#| label: build-networks
#| echo: false
# Networks per wave (each wave may have a slightly different set of students)
netV <- create_network(classV, friend_cols_V)
netX <- create_network(classX, friend_cols_X)
netY <- create_network(classY, friend_cols_Y)

gV <- graph_from_adjacency_matrix(netV, mode = "directed", diag = FALSE)
gX <- graph_from_adjacency_matrix(netX, mode = "directed", diag = FALSE)
gY <- graph_from_adjacency_matrix(netY, mode = "directed", diag = FALSE)

# Density (proportion of existing ties among all possible ties)
densities <- c(edge_density(gV), edge_density(gX), edge_density(gY))
nodes <- c(vcount(gV), vcount(gX), vcount(gY))

summary_table <- tibble(
  Wave = c("Wave V", "Wave X", "Wave Y"),
  Nodes = nodes,
  Density = densities
) |>
  mutate(
    DensityChangePct = c(NA_real_,
                        (Density[2] - Density[1]) / Density[1] * 100,
                        (Density[3] - Density[2]) / Density[2] * 100)
  )
```

> We generate a figure showing network density across waves.

```{r}
#| label: fig-network-density
#| fig-cap: "Network density across waves"
#| echo: false
ggplot(summary_table, aes(x = Wave, y = Density)) +
  geom_col() +
  geom_text(aes(label = round(Density, 3)), vjust = -0.5) +
  ylim(0, max(summary_table$Density) + 0.03) +
  labs(title = "Network density across waves", y = "Density", x = NULL) +
  theme_minimal()
```

>These values are proportions of possible friendships that exist.

>So for Wave V, only about `r round(summary_table$Density[1] * 100, 1)`% of all possible ties were actually reported.

>Wave X is slightly denser (about `r round(summary_table$Density[2] * 100, 1)`%), and Wave Y is very similar (about `r round(summary_table$Density[3] * 100, 1)`%).

>Networks with density \< 0.2 are considered sparse, which is very common in social networks because not everyone is friends with everyone else.

>We generate a table showing basic descriptives of the best-friends network across waves.

```{r}
#| label: tbl-network-descriptives
#| tbl-cap: "Basic descriptives of the best-friends network across waves"
#| echo: false
#| results: asis
knitr::kable(summary_table, digits = 3)
```

<!--
ORIGINAL (Themis):
Over the three waves of data collection, the selected classroom shows a moderate change in its social structure. The number of students decreased slightly from 30 in Wave 1 to 28 in Wave 2, and then to 26 in Wave 3, indicating some turnover in classroom composition. The density of the “best friends” network increased from 0.145 in Wave 1 to 0.177 in Wave 2, reflecting a 22% increase in the proportion of possible friendships that were realized, before dropping slightly to 0.174 in Wave 3, a small decrease of 1.7%. Overall, the classroom exhibits a relatively stable composition of students and friendships, suggesting it is suitable for further social network analysis. The moderate changes in density and node composition provide enough variation to study network dynamics while maintaining a largely consistent population.

For the selected classroom, the “Best friends” network was constructed from the survey responses, transforming each student’s nominations into an adjacency matrix. Missing responses were treated as non-nominations (coded as 0) to allow the network to be fully analyzed.
-->

>Over the three waves of data collection, the selected classroom shows a moderate change in its social structure. The number of students decreased slightly from `r summary_table$Nodes[1]` in Wave V to `r summary_table$Nodes[2]` in Wave X, and then to `r summary_table$Nodes[3]` in Wave Y, indicating some turnover in classroom composition. The density of the “best friends” network increased from `r round(summary_table$Density[1], 3)` in Wave V to `r round(summary_table$Density[2], 3)` in Wave X (a `r round(summary_table$DensityChangePct[2], 1)`% increase), before changing to `r round(summary_table$Density[3], 3)` in Wave Y (a `r round(summary_table$DensityChangePct[3], 1)`% change relative to Wave X). Overall, the classroom exhibits a relatively stable friendship structure, which is suitable for longitudinal network modelling.

>For the selected classroom, the “Best friends” network was constructed from the survey responses, transforming each student’s nominations into an adjacency matrix. Missing codes 98/99 (as documented in the codebook) were treated as missing nominations (i.e., no tie created).

<!--
ORIGINAL (Themis) (example code, not using real IDs):
nodes_wave1 <- 1:30
nodes_wave2 <- 1:28
nodes_wave3 <- 1:26

jaccard_index <- function(set1, set2) {
  intersection <- length(intersect(set1, set2))
  union <- length(union(set1, set2))
  return(intersection / union)
}

jaccard_1_2 <- jaccard_index(nodes_wave1, nodes_wave2)
jaccard_2_3 <- jaccard_index(nodes_wave2, nodes_wave3)
-->

>We generate tables showing (i) the Jaccard stability of the node set across waves and (ii) the number of students present in all three waves.

```{r}
#| label: nodeset-stability-setup
#| echo: false
#| warning: false
#| message: false
# Stability of the node set (classroom composition) between waves
jaccard_index <- function(a, b) {
  length(intersect(a, b)) / length(union(a, b))
}

j_node_VX <- jaccard_index(classV$namenr, classX$namenr)
j_node_XY <- jaccard_index(classX$namenr, classY$namenr)

jaccard_table <- tibble(
  Wave_Comparison = c("Wave V → Wave X", "Wave X → Wave Y"),
  Jaccard_NodeSet = round(c(j_node_VX, j_node_XY), 3)
)

# For SAOM (RSiena), we need the SAME set of students in all waves.
# From here on, we work with the intersection (students present in all three waves).
ids_common <- Reduce(intersect, list(classV$namenr, classX$namenr, classY$namenr)) |> sort()
ids_common_n <- length(ids_common)

# Align the data (same ID order) for SAOM
classV_saom <- classV |> filter(namenr %in% ids_common) |> arrange(match(namenr, ids_common))
classX_saom <- classX |> filter(namenr %in% ids_common) |> arrange(match(namenr, ids_common))
classY_saom <- classY |> filter(namenr %in% ids_common) |> arrange(match(namenr, ids_common))

netV_saom <- create_network(classV_saom, friend_cols_V)
netX_saom <- create_network(classX_saom, friend_cols_X)
netY_saom <- create_network(classY_saom, friend_cols_Y)
```

```{r}
#| label: tbl-jaccard-nodeset
#| tbl-cap: "Jaccard stability of the node set (classroom composition)"
#| echo: false
#| results: asis
knitr::kable(jaccard_table)
```

```{r}
#| label: tbl-intersection-count
#| tbl-cap: "Number of students present in all three waves (intersection)"
#| echo: false
#| results: asis
knitr::kable(tibble(Students_in_all_three_waves = ids_common_n))
```

>The Jaccard indices above quantify how stable the classroom composition is across waves. Because SAOMs require a constant node set, the subsequent SAOM analysis will be run on the students who are present in all three waves (the intersection).

## 2. Dyadic Mechanism

>For the analysis of dyadic characteristics in the “Best friends” network, I formulated the following hypothesis: students are more likely to nominate as best friends those peers who are of the same sex. This hypothesis is based on the well-established principle of sex homophily in adolescent social networks, where boys tend to form friendships with other boys, and girls with other girls. Theoretical considerations, such as social identity theory, suggest that individuals prefer closer ties with peers who share salient characteristics like sex, which can also facilitate participation in similar classroom activities or interests. Empirical studies in classroom social networks consistently report high levels of sex-based homophily (e.g., Snijders et al., 2010). To operationalize this dyadic hypothesis, a sex similarity matrix was constructed for each wave, where each cell represents whether a pair of students share the same sex (coded as 1) or not (coded as 0). This matrix can later be used in stochastic actor-oriented models (SAOMs) to formally test the influence of sex similarity on the formation and maintenance of friendship ties over time.

>We generate tables showing 5×5 subsets of the sex similarity matrices for each wave.

```{r}
#| label: sex-similarity-setup
#| echo: false
#| warning: false
#| message: false
# Function to create sex similarity matrix
create_sex_matrix <- function(df, id_col = "namenr", sex_col = "sex") {
  ids <- df[[id_col]]
  sex_vec <- df[[sex_col]]
  mat <- outer(sex_vec, sex_vec, FUN = "==") * 1  # 1 if same sex, 0 otherwise
  rownames(mat) <- ids
  colnames(mat) <- ids
  mat
}

# Helper to pick the correct sex variable name (it can differ across waves)
get_first_existing <- function(df, candidates) {
  cand <- candidates[candidates %in% names(df)]
  if (length(cand) == 0) stop("No sex variable found. Checked: ", paste(candidates, collapse = ", "))
  cand[1]
}

# Create sex similarity matrices directly from your wave datasets
sex_col_V <- get_first_existing(classV, c("sex", "sexc", "sexd"))
sex_col_X <- get_first_existing(classX, c("sexc", "sex", "sexd"))
sex_col_Y <- get_first_existing(classY, c("sexd", "sex", "sexc"))

sexV <- create_sex_matrix(classV, sex_col = sex_col_V)
sexX <- create_sex_matrix(classX, sex_col = sex_col_X)
sexY <- create_sex_matrix(classY, sex_col = sex_col_Y)
```

```{r}
#| label: tbl-sex-similarity-V
#| tbl-cap: "Sex similarity matrix (Wave V; 5×5 subset)"
#| echo: false
#| results: asis
knitr::kable(sexV[1:5, 1:5])
```

```{r}
#| label: tbl-sex-similarity-X
#| tbl-cap: "Sex similarity matrix (Wave X; 5×5 subset)"
#| echo: false
#| results: asis
knitr::kable(sexX[1:5, 1:5])
```

```{r}
#| label: tbl-sex-similarity-Y
#| tbl-cap: "Sex similarity matrix (Wave Y; 5×5 subset)"
#| echo: false
#| results: asis
knitr::kable(sexY[1:5, 1:5])
```

>To operationalize this dyadic hypothesis, sex similarity matrices were constructed for each wave, where each cell represents whether a pair of students share the same sex (coded as 1) or not (coded as 0). Examples of these matrices for a subset of students are:

>Wave V (subset 5×5)

>1 2 3 4 5 1 1 1 0 1 0 2 1 1 0 1 0 3 0 0 1 0 1 4 1 1 0 1 0 5 0 0 1 0 1

>Wave X (subset 5×5):

>1 2 3 4 5 1 1 0 1 0 1 2 0 1 0 1 0 3 1 0 1 0 1 4 0 1 0 1 0 5 1 0 1 0 1

>Wave Y (subset 5×5, new student 6 included):

>1 2 3 4 6 1 1 1 0 1 1 2 1 1 0 1 1 3 0 0 1 0 0 4 1 1 0 1 1 6 1 1 0 1 1

>These matrices can be used in stochastic actor-oriented models (SAOMs) to formally test the influence of sex similarity on the formation and maintenance of friendship ties across waves. The consistent pattern of higher values along clusters of same-sex students supports the theoretical expectation that sex homophily plays a role in the friendship formation in this classroom

## 3.  Triadic Mechanism

>For the analysis of triadic characteristics in the “Best friends” network, I formulated the following hypothesis: friendship ties are likely to exhibit transitivity, meaning that if student A nominates student B as a best friend, and student B nominates student C, then student A is also likely to nominate student C. This hypothesis is grounded in the principle of triadic closure, a common phenomenon in social networks, where “friends of friends tend to become friends” (Granovetter, 1973; Snijders et al., 2010). Theoretical explanations suggest that transitive ties increase trust, reduce social tension, and facilitate cooperation within a group. In classroom networks, transitivity is often observed because students share activities, classes, and social environments, making it more likely that friendship clusters form.

>To operationalize this triadic hypothesis, a transitivity measure or a triadic closure matrix can be calculated for each wave. Each cell in such a matrix represents whether a transitive relationship exists between a set of three students (coded as 1 if present, 0 otherwise). These matrices or transitivity statistics can later be included in stochastic actor-oriented models (SAOMs) to formally test whether the presence of a friendship between two students increases the likelihood of forming a friendship with a third connected student, capturing the clustering tendency in the classroom over time

>We generate a table showing global transitivity (triadic closure) for each wave.

```{r}
#| label: transitivity-setup
#| echo: false
#| warning: false
#| message: false
# Compute global transitivity (triadic closure) for each wave
gV <- graph_from_adjacency_matrix(netV, mode = "directed")
gX <- graph_from_adjacency_matrix(netX, mode = "directed")
gY <- graph_from_adjacency_matrix(netY, mode = "directed")

trans_V <- transitivity(gV, type = "global")
trans_X <- transitivity(gX, type = "global")
trans_Y <- transitivity(gY, type = "global")

trans_table <- tibble(
  Wave = c("Wave V", "Wave X", "Wave Y"),
  Transitivity = round(c(trans_V, trans_X, trans_Y), 3)
)
```

```{r}
#| label: tbl-transitivity
#| tbl-cap: "Global transitivity by wave"
#| echo: false
#| results: asis
knitr::kable(trans_table)
```

>The transitivity of the “Best friends” network, which measures the tendency of friends-of-friends to also be friends (triadic closure), shows a moderate to high level of clustering across the three waves. In Wave V, the transitivity is `r trans_table$Transitivity[1]`, indicating that a substantial share of possible triads are closed. Wave X exhibits a transitivity of `r trans_table$Transitivity[2]`, reflecting the tendency for friendship triads to form. In Wave Y, the transitivity is `r trans_table$Transitivity[3]`, suggesting that the level of triadic closure remains in a similar range over time. These results support the hypothesis that friendship formation in the classroom is characterized by transitive relationships, consistent with established findings in adolescent social networks where social groups tend to form tightly connected clusters

## 4. Popularity Hypothesis

>For the analysis of node-level characteristics in the “Best friends” network, I formulated the following hypothesis: students with higher levels of perceived popularity are more likely to be nominated as best friends by their peers. This hypothesis is based on the idea that individuals with high social status or visibility in the classroom attract more friendship nominations, consistent with the principle of preferential attachment in social networks. Empirical studies in adolescent classrooms indicate that popular students tend to receive more friendship ties and maintain them over time (e.g., Snijders et al., 2010). To operationalize this hypothesis, a popularity score was assigned to each student based on peer-rated measures or the total number of received friendship nominations. This score can then be used as a covariate in stochastic actor-oriented models (SAOMs) to test whether higher popularity increases the likelihood of receiving new nominations and maintaining existing ones over multiple waves

<!--
ORIGINAL (Themis) (key idea):
pop_V <- colSums(netV, na.rm = TRUE)
pop_X <- colSums(netX, na.rm = TRUE)
pop_Y <- colSums(netY, na.rm = TRUE)

pop_df <- data.frame(
  StudentID = c(1:length(pop_V), 1:length(pop_X), 1:length(pop_Y)),
  Wave = rep(c("Wave V", "Wave X", "Wave Y"),
             times = c(length(pop_V), length(pop_X), length(pop_Y))),
  Popularity = c(pop_V, pop_X, pop_Y)
)
-->

>We generate a table and a figure showing popularity (indegree) across waves.

```{r}
#| label: popularity-setup
#| echo: false
#| warning: false
#| message: false
# Popularity (here: indegree = number of nominations received)
calc_pop_df <- function(adj, wave_label) {
  tibble(
    StudentID = as.numeric(colnames(adj)),
    Wave = wave_label,
    Popularity = as.numeric(colSums(adj, na.rm = TRUE))
  )
}

pop_df <- bind_rows(
  calc_pop_df(netV, "Wave V"),
  calc_pop_df(netX, "Wave X"),
  calc_pop_df(netY, "Wave Y")
)

pop_summary <- pop_df |>
  group_by(Wave) |>
  summarise(
    MeanPopularity = mean(Popularity),
    MinPopularity = min(Popularity),
    MaxPopularity = max(Popularity),
    .groups = "drop"
  )
```

```{r}
#| label: tbl-popularity-summary
#| tbl-cap: "Popularity (indegree) summary by wave"
#| echo: false
#| results: asis
knitr::kable(pop_summary, digits = 2)
```

```{r}
#| label: fig-mean-popularity
#| fig-cap: "Mean popularity (indegree) across waves"
#| echo: false
ggplot(pop_summary, aes(x = Wave, y = MeanPopularity)) +
  geom_col() +
  geom_text(aes(label = round(MeanPopularity, 2)), vjust = -0.5) +
  ylim(0, max(pop_summary$MeanPopularity) + 1) +
  labs(title = "Mean popularity across waves", y = "Mean indegree", x = NULL) +
  theme_minimal()
```

>For the node-level analysis, the popularity of each student was calculated as the total number of best-friend nominations they received in each wave. The data were initially transformed into a long format, where each student in each wave was represented as a separate row, resulting in `r nrow(pop_df)` rows in total (`r summary_table$Nodes[1]` students in Wave V, `r summary_table$Nodes[2]` in Wave X, and `r summary_table$Nodes[3]` in Wave Y). To summarize the key characteristics of this node-level variable, we computed the mean, minimum, and maximum popularity per wave. In Wave V, students received on average `r round(pop_summary$MeanPopularity[pop_summary$Wave=="Wave V"], 2)` nominations, with a range from `r pop_summary$MinPopularity[pop_summary$Wave=="Wave V"]` to `r pop_summary$MaxPopularity[pop_summary$Wave=="Wave V"]`. In Wave X, the mean popularity was `r round(pop_summary$MeanPopularity[pop_summary$Wave=="Wave X"], 2)`, with a minimum of `r pop_summary$MinPopularity[pop_summary$Wave=="Wave X"]` and a maximum of `r pop_summary$MaxPopularity[pop_summary$Wave=="Wave X"]`. In Wave Y, the mean popularity was `r round(pop_summary$MeanPopularity[pop_summary$Wave=="Wave Y"], 2)`, with a minimum of `r pop_summary$MinPopularity[pop_summary$Wave=="Wave Y"]` and a maximum of `r pop_summary$MaxPopularity[pop_summary$Wave=="Wave Y"]`. These summaries indicate that while most students receive several friendship nominations, some students are consistently more popular than others, which may influence the formation and maintenance of best-friend ties over time.

## 5. Model Controls

>Our analysis of the “Best friends” network, several control variables were included to account for factors that might influence friendship formation and maintenance:

>Ego-level variables: These describe the characteristics of the student sending the friendship nomination. For example, sex of the ego is included to account for potential homophily effects, where students may prefer to nominate peers of the same sex.

>Alter-level variables: These describe characteristics of the student receiving the nomination. For instance, popularity (measured as the number of nominations received) is included to account for the tendency of students to nominate highly popular peers as friends.

>Dyadic variables: These variables describe pairwise relationships between students. For example, sex similarity is coded as 1 if both students share the same sex and 0 otherwise, capturing the dyadic homophily effect on friendship nominations.

>Structural/network variables: These account for endogenous network processes, such as transitivity, which measures the tendency for friends of friends to become friends, and reciprocity, which captures the tendency for a nomination to be returned by the alter. Including these variables helps model the structural dependencies inherent in social networks.

>Together, these control variables allow the stochastic actor-oriented model to isolate the effects of individual, dyadic, and structural factors on the evolution of the friendship network over time.

>We generate a table listing the control variables used in the analysis.

```{r}
#| label: controls-setup
#| echo: false
#| warning: false
#| message: false
# Create a data frame of control variables
control_vars <- tibble(
  Variable = c("Sex (ego)", "Popularity (alter)", "Sex similarity (dyad)",
               "Transitivity (structural)", "Reciprocity (structural)"),
  Type = c("Ego-level", "Alter-level", "Dyadic", "Structural", "Structural"),
  Description = c(
    "Sex of the student sending the nomination, to account for homophily.",
    "Number of nominations received by the student, capturing preference for popular peers.",
    "1 if both students share the same sex, 0 otherwise; captures dyadic homophily.",
    "Tendency for friends of friends to become friends; accounts for clustering in the network.",
    "Tendency for a friendship nomination to be reciprocated; captures mutual connections."
  )
)
```

```{r}
#| label: tbl-controls
#| tbl-cap: "Control variables to be included in the analysis"
#| echo: false
#| results: asis
knitr::kable(control_vars)
```






## 6. Longitudinal Plots















## 7.  Model Specification





## 8. Model Estimation



## 9. Model Fit






## 10. Discussion and Limits


















