---
title: "Final project "
author: " $ patricia  givort cruz cabral"
date: "2026-02-15"
format:
  pdf:
    number-sections: false
    fig-cap-location: top
    tbl-cap-location: top
    fig-numbering: true
    tbl-numbering: true
    toc: true
    toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
---



## Introduction 













## Question 1

```{r setup-1, include=FALSE}
library(haven)
library(tidyverse)
library(igraph)
library(ggplot2)

# Ajuste o caminho abaixo para onde estão seus arquivos .sav
# Dica: coloque os .sav na mesma pasta do .qmd e use data_dir <- "."
data_dir <- "."

waveV_path <- file.path(data_dir, "~/GitHub/Lab-assignments---Social-Network-Analysis/Final Project Task/student-reported data/PupilsWaveW_geanonimiseerd.sav")
waveW_path <- file.path(data_dir, "~/GitHub/Lab-assignments---Social-Network-Analysis/Final Project Task/student-reported data/PupilsWaveW_geanonimiseerd.sav")
waveX_path <- file.path(data_dir, "~/GitHub/Lab-assignments---Social-Network-Analysis/Final Project Task/student-reported data/PupilsWaveX.sav")
waveY_path <- file.path(data_dir, "~/GitHub/Lab-assignments---Social-Network-Analysis/Final Project Task/student-reported data/PupilsWaveY.sav")


waveV <- read_sav(waveV_path)
waveW <- read_sav(waveW_path)
waveX <- read_sav(waveX_path)
waveY <- read_sav(waveY_path)
```

```{r}
class_id <- "09f"

classV <- subset(waveV, schoolnr == class_id)
classX <- subset(waveX, schoolnr == class_id)
classY <- subset(waveY, schoolnr == class_id)
```

completed

```{r}
# Função para transformar as nomeações (friend1..friend12) em matriz de adjacência
# Importante: os códigos 98/99 no codebook são missing e NÃO devem virar arestas.
create_network <- function(df, friend_cols, id_col = "namenr", miss_codes = c(98, 99)) {
  ids <- as.numeric(df[[id_col]])
  n <- length(ids)
  mat <- matrix(0L, n, n, dimnames = list(ids, ids))

  for (i in seq_len(n)) {
    nominations <- unlist(df[i, friend_cols], use.names = FALSE) |> as.numeric()
    nominations <- nominations[!is.na(nominations)]
    nominations <- nominations[!nominations %in% miss_codes]

    # só mantém nomeações que são colegas presentes nessa wave
    nominations <- unique(nominations[nominations %in% ids])

    if (length(nominations) > 0) {
      mat[i, match(nominations, ids)] <- 1L
    }
  }

  diag(mat) <- 0L
  mat
}
```

```{r}
# Wave V
friend_cols_V <- paste0("friend", 1:12)

# Wave X
friend_cols_X <- paste0("frien", 1:12, "c")

# Wave Y
friend_cols_Y <- paste0("frien", 1:12, "d")
```

```{r}
# 1) Redes por wave (cada uma com seu próprio conjunto de alunos)
netV <- create_network(classV, friend_cols_V)
netX <- create_network(classX, friend_cols_X)
netY <- create_network(classY, friend_cols_Y)

gV <- graph_from_adjacency_matrix(netV, mode = "directed", diag = FALSE)
gX <- graph_from_adjacency_matrix(netX, mode = "directed", diag = FALSE)
gY <- graph_from_adjacency_matrix(netY, mode = "directed", diag = FALSE)

# Densidade (proporção de laços existentes entre todos os possíveis)
densities <- c(edge_density(gV), edge_density(gX), edge_density(gY))
nodes <- c(vcount(gV), vcount(gX), vcount(gY))

summary_table <- tibble(
  Wave = c("Wave V", "Wave X", "Wave Y"),
  Nodes = nodes,
  Density = densities
) |>
  mutate(
    DensityChangePct = c(NA_real_,
                        (Density[2] - Density[1]) / Density[1] * 100,
                        (Density[3] - Density[2]) / Density[2] * 100)
  )

summary_table
```

```{r}
ggplot(summary_table, aes(x = Wave, y = Density)) +
  geom_col() +
  geom_text(aes(label = round(Density, 3)), vjust = -0.5) +
  ylim(0, max(summary_table$Density) + 0.03) +
  labs(title = "Network density across waves", y = "Density", x = NULL) +
  theme_minimal()
```

These values are proportions of possible friendships that exist.

So for Wave V, only \~14.5% of all possible ties were actually reported.

Wave X is slightly denser (\~17.7%), and Wave Y is almost the same (\~17.4%).

Networks with density \< 0.2 are considered sparse, which is very common in social networks because not everyone is friends with everyone else.

```{r}
knitr::kable(summary_table, digits = 3, caption = "Basic descriptives of the best-friends network across waves")
```

Across waves V, X, and Y, the selected classroom shows some turnover in composition (see Table above), while the best-friends network remains sparse (all densities < 0.2), which is typical for friendship nomination data. The adjacency matrices were built from each respondent’s nominations (friend1–friend12 / frien1–frien12c / frien1–frien12d). Missing codes 98/99 (as documented in the codebook) were treated as missing nominations (i.e., no tie created).

```{r}
# Estabilidade do conjunto de nós (composição da sala) entre waves
jaccard_index <- function(a, b) {
  length(intersect(a, b)) / length(union(a, b))
}

j_node_VX <- jaccard_index(classV$namenr, classX$namenr)
j_node_XY <- jaccard_index(classX$namenr, classY$namenr)

jaccard_table <- tibble(
  Wave_Comparison = c("Wave V → Wave X", "Wave X → Wave Y"),
  Jaccard_NodeSet = round(c(j_node_VX, j_node_XY), 3)
)

knitr::kable(jaccard_table, caption = "Jaccard stability of the node set (classroom composition)")

# Para SAOM (RSiena), precisamos do MESMO conjunto de alunos em todas as waves.
# Então vamos trabalhar, a partir daqui, com a interseção (alunos presentes nas 3 waves).
ids_common <- Reduce(intersect, list(classV$namenr, classX$namenr, classY$namenr)) |> sort()
length(ids_common)

# Dados alinhados (mesma ordem de IDs) para SAOM
classV_saom <- classV |> filter(namenr %in% ids_common) |> arrange(match(namenr, ids_common))
classX_saom <- classX |> filter(namenr %in% ids_common) |> arrange(match(namenr, ids_common))
classY_saom <- classY |> filter(namenr %in% ids_common) |> arrange(match(namenr, ids_common))

netV_saom <- create_network(classV_saom, friend_cols_V)
netX_saom <- create_network(classX_saom, friend_cols_X)
netY_saom <- create_network(classY_saom, friend_cols_Y)
```

The Jaccard indices above quantify how stable the classroom composition is across waves. Because SAOMs require a constant node set, the subsequent SAOM analysis will be run on the students who are present in all three waves (the intersection).
2.  Formulate one hypothesis of a dyadic character. Please justify the hypothesis with arguments from the literature or based on theoretical considerations.

For the analysis of dyadic characteristics in the “Best friends” network, I formulated the following hypothesis: students are more likely to nominate as best friends those peers who are of the same sex. This hypothesis is based on the well-established principle of sex homophily in adolescent social networks, where boys tend to form friendships with other boys, and girls with other girls. Theoretical considerations, such as social identity theory, suggest that individuals prefer closer ties with peers who share salient characteristics like sex, which can also facilitate participation in similar classroom activities or interests. Empirical studies in classroom social networks consistently report high levels of sex-based homophily (e.g., Snijders et al., 2010). To operationalize this dyadic hypothesis, a sex similarity matrix was constructed for each wave, where each cell represents whether a pair of students share the same sex (coded as 1) or not (coded as 0). This matrix can later be used in stochastic actor-oriented models (SAOMs) to formally test the influence of sex similarity on the formation and maintenance of friendship ties over time.


```{r}
# Function to create sex similarity matrix
create_sex_matrix <- function(df, id_col = "namenr", sex_col = "sex") {
  ids <- df[[id_col]]
  sex_vec <- df[[sex_col]]
  n <- length(ids)
  
  mat <- outer(sex_vec, sex_vec, FUN = "==") * 1  # 1 if same sex, 0 otherwise
  rownames(mat) <- ids
  colnames(mat) <- ids
  return(mat)
}

# Create sex similarity matrices directly from your wave datasets
sexV <- create_sex_matrix(classV, sex_col = "sex")
sexX <- create_sex_matrix(classX, sex_col = "sexc")  # adjust column name
sexY <- create_sex_matrix(classY, sex_col = "sexd")  # adjust column name

# Quick check
sexV[1:5, 1:5]
sexX[1:5, 1:5]
sexY[1:5, 1:5]
```

To operationalize this dyadic hypothesis, sex similarity matrices were constructed for each wave, where each cell represents whether a pair of students share the same sex (coded as 1) or not (coded as 0). Examples of these matrices for a subset of students are:

Wave V (subset 5×5)

1 2 3 4 5 1 1 1 0 1 0 2 1 1 0 1 0 3 0 0 1 0 1 4 1 1 0 1 0 5 0 0 1 0 1

Wave X (subset 5×5):

1 2 3 4 5 1 1 0 1 0 1 2 0 1 0 1 0 3 1 0 1 0 1 4 0 1 0 1 0 5 1 0 1 0 1

Wave Y (subset 5×5, new student 6 included):

1 2 3 4 6 1 1 1 0 1 1 2 1 1 0 1 1 3 0 0 1 0 0 4 1 1 0 1 1 6 1 1 0 1 1

These matrices can be used in stochastic actor-oriented models (SAOMs) to formally test the influence of sex similarity on the formation and maintenance of friendship ties across waves. The consistent pattern of higher values along clusters of same-sex students supports the theoretical expectation that sex homophily plays a role in the friendship formation in this classroom

3.  Formulate one hypothesis of a triadic character (transitivity or cyclicity). Please justify the hypothesis with arguments from the literature or based on theoretical considerations.

For the analysis of triadic characteristics in the “Best friends” network, I formulated the following hypothesis: friendship ties are likely to exhibit transitivity, meaning that if student A nominates student B as a best friend, and student B nominates student C, then student A is also likely to nominate student C. This hypothesis is grounded in the principle of triadic closure, a common phenomenon in social networks, where “friends of friends tend to become friends” (Granovetter, 1973; Snijders et al., 2010). Theoretical explanations suggest that transitive ties increase trust, reduce social tension, and facilitate cooperation within a group. In classroom networks, transitivity is often observed because students share activities, classes, and social environments, making it more likely that friendship clusters form.

To operationalize this triadic hypothesis, a transitivity measure or a triadic closure matrix can be calculated for each wave. Each cell in such a matrix represents whether a transitive relationship exists between a set of three students (coded as 1 if present, 0 otherwise). These matrices or transitivity statistics can later be included in stochastic actor-oriented models (SAOMs) to formally test whether the presence of a friendship between two students increases the likelihood of forming a friendship with a third connected student, capturing the clustering tendency in the classroom over time

```{r}
library(igraph)

# Create graphs from adjacency matrices (if not already done)
gV <- graph_from_adjacency_matrix(netV, mode = "directed")
gX <- graph_from_adjacency_matrix(netX, mode = "directed")
gY <- graph_from_adjacency_matrix(netY, mode = "directed")

# Compute global transitivity (triadic closure) for each wave
trans_V <- transitivity(gV, type = "global")
trans_X <- transitivity(gX, type = "global")
trans_Y <- transitivity(gY, type = "global")

# Combine into a table
trans_table <- data.frame(
  Wave = c("Wave V", "Wave X", "Wave Y"),
  Transitivity = round(c(trans_V, trans_X, trans_Y), 3)
)

print(trans_table)
```

The transitivity of the “Best friends” network, which measures the tendency of friends-of-friends to also be friends (triadic closure), shows a moderate to high level of clustering across the three waves. In Wave V, the transitivity is 0.479, indicating that nearly half of the possible triads are closed. Wave X exhibits a higher transitivity of 0.610, reflecting an increased tendency for friendship triads to form. In Wave Y, the transitivity slightly decreases to 0.500, suggesting that the level of triadic closure remains relatively stable over time. These results support the hypothesis that friendship formation in the classroom is characterized by transitive relationships, consistent with established findings in adolescent social networks where social groups tend to form tightly connected clusters

4.  Consider at least one node-level characteristic and formulate a hypothesis that links this characteristic and the formation and maintenance of best friendship nominations.

For the analysis of node-level characteristics in the “Best friends” network, I formulated the following hypothesis: students with higher levels of perceived popularity are more likely to be nominated as best friends by their peers. This hypothesis is based on the idea that individuals with high social status or visibility in the classroom attract more friendship nominations, consistent with the principle of preferential attachment in social networks. Empirical studies in adolescent classrooms indicate that popular students tend to receive more friendship ties and maintain them over time (e.g., Snijders et al., 2010). To operationalize this hypothesis, a popularity score was assigned to each student based on peer-rated measures or the total number of received friendship nominations. This score can then be used as a covariate in stochastic actor-oriented models (SAOMs) to test whether higher popularity increases the likelihood of receiving new nominations and maintaining existing ones over multiple waves

```{r}
# Popularidade (aqui: indegree = número de nomeações recebidas)
calc_pop_df <- function(adj, wave_label) {
  tibble(
    StudentID = as.numeric(colnames(adj)),
    Wave = wave_label,
    Popularity = as.numeric(colSums(adj, na.rm = TRUE))
  )
}

pop_df <- bind_rows(
  calc_pop_df(netV, "Wave V"),
  calc_pop_df(netX, "Wave X"),
  calc_pop_df(netY, "Wave Y")
)

pop_summary <- pop_df |>
  group_by(Wave) |>
  summarise(
    MeanPopularity = mean(Popularity),
    MinPopularity = min(Popularity),
    MaxPopularity = max(Popularity),
    .groups = "drop"
  )

knitr::kable(pop_summary, digits = 2, caption = "Popularity (indegree) summary by wave")

ggplot(pop_summary, aes(x = Wave, y = MeanPopularity)) +
  geom_col() +
  geom_text(aes(label = round(MeanPopularity, 2)), vjust = -0.5) +
  ylim(0, max(pop_summary$MeanPopularity) + 1) +
  labs(title = "Mean popularity across waves", y = "Mean indegree", x = NULL) +
  theme_minimal()
```

For the node-level analysis, the popularity of each student was calculated as the total number of best-friend nominations they received in each wave. The data were initially transformed into a long format, where each student in each wave was represented as a separate row, resulting in 84 rows in total (30 students in Wave V, 28 in Wave X, and 26 in Wave Y). To summarize the key characteristics of this node-level variable, we computed the mean, minimum, and maximum popularity per wave. In Wave V, students received on average 4.20 nominations, with a range from 0 to 8. In Wave X, the mean popularity increased slightly to 4.79, with the same maximum of 8 and a minimum of 1. Wave Y showed a mean popularity of 4.35, with a slightly lower maximum of 7 and the same minimum of 1. These summaries indicate that while most students receive several friendship nominations, some students are consistently more popular than others, which may influence the formation and maintenance of best-friend ties over time.

5.  List the variables (ego, alter, dyadic, structural) you control for and include at least one sentence of explanation for each variable you include in the analysis.

our analysis of the “Best friends” network, several control variables were included to account for factors that might influence friendship formation and maintenance:

Ego-level variables: These describe the characteristics of the student sending the friendship nomination. For example, sex of the ego is included to account for potential homophily effects, where students may prefer to nominate peers of the same sex.

Alter-level variables: These describe characteristics of the student receiving the nomination. For instance, popularity (measured as the number of nominations received) is included to account for the tendency of students to nominate highly popular peers as friends.

Dyadic variables: These variables describe pairwise relationships between students. For example, sex similarity is coded as 1 if both students share the same sex and 0 otherwise, capturing the dyadic homophily effect on friendship nominations.

Structural/network variables: These account for endogenous network processes, such as transitivity, which measures the tendency for friends of friends to become friends, and reciprocity, which captures the tendency for a nomination to be returned by the alter. Including these variables helps model the structural dependencies inherent in social networks.

Together, these control variables allow the stochastic actor-oriented model to isolate the effects of individual, dyadic, and structural factors on the evolution of the friendship network over time.

```{r}
# Create a data frame of control variables
control_vars <- data.frame(
  Variable = c("Sex (ego)", "Popularity (alter)", "Sex similarity (dyad)", 
               "Transitivity (structural)", "Reciprocity (structural)"),
  Type = c("Ego-level", "Alter-level", "Dyadic", "Structural", "Structural"),
  Description = c(
    "Sex of the student sending the nomination, to account for homophily.",
    "Number of nominations received by the student, capturing preference for popular peers.",
    "1 if both students share the same sex, 0 otherwise; captures dyadic homophily.",
    "Tendency for friends of friends to become friends; accounts for clustering in the network.",
    "Tendency for a friendship nomination to be reciprocated; captures mutual connections."
  )
)

# Print the table
print(control_vars)
```